---
title: "R Basics Exam Review"
tutorial:
  id: "Week_6_Exam_Review"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
description: "Practice problems for Exam II focused on the basics of R"
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
gradethis::gradethis_setup()
options(tutorial.event_recorder = tutorialize::recorder)
options(tutorial.storage = learnr::filesystem_storage(file.path(path.expand("~"), "_learnr_storage_")))
```

## Key Concepts to Review for Exam 1

* *Atomic Vectors*, be able to define, identify, and apply them.
* *Recycling*, be able to define, identify, and apply this concept to data.
* *Operators*, including `:`, `<-`, `$`, `[]`, and all arithmetic and comparison operators.
* *Functions*, including `c()`,`plot()`,`seq()`,`head()`,`str()`,`mean()`, and others covered through the extraction tutorial. *Be able to write a simple custom function.*
* *Dataframes*, including what they are, the three tidy principles, how to make them, and how to manipulate them.
* *Extraction*, be able to define, identify, and apply this concept to data using operators.

*In addition to these practice problems, the best way to study for the exam is to review each tutorial and lab assigned throughout this section.*

## Vectors Practice Problems

Q1: Use the `:` operator to create a numeric vector from 1 to 10 that increases by values of 1. Name this vector "vector_1"

```{r problem_vector, exercise=tutorialize}
vector_1 <- 1:10
```

Q2: Use the `c()` function to create a character vector that spells out BIOME, with each value being its own letter. Name this vector "biome".

```{r problem_vector_2, exercise=tutorialize}
biome <- c("B","I","O","M","E")
```

Q3: Use recycling to multiply vector_1 by 0.75.

```{r problem_vector_3, exercise=tutorialize,exercise.setup=problem_vector}
vector_1*0.75
```

Q4: Create a new vector using the `seq()` function which ranges from 10 to 80 and increases by values of 10. Name this new vector vector_2. Then, use the `plot()` function to plot vector_1 on the x axis and vector_2 on the y axis.

```{r problem_vector_4, exercise=tutorialize,exercise.setup=problem_vector}
vector_2 <- seq(from = 8,to = 80,by = 10)
plot(vector_1,vector_2)
```

```{r quiz_vectors}

learnr:: quiz( 
  question("What is the purpose of the `<-` operator?",
    answer("to prevent a line of code from running"),
    answer("add two values"),
    answer("compare two values"),
    answer("assign a name to a value", correct = TRUE)
  ),
   question("What is the purpose of the `:` operator?",
    answer("indicate complete equality"),
    answer("create a sequence of numbers",correct = TRUE),
    answer("compare two values"),
    answer("extract a value from a sequence")
  ),
  question("True or False: all values in an atomic vector are composed of the same data type",
    answer("True",correct = TRUE),
    answer("False")
  ),
  question("What does the c() function do?",
    answer("prevents a line of code from running"),
    answer("adds two vectors together"),
    answer("combines values in a sequence to create a vector",correct = TRUE),
    answer("creates a sequence of values according to a range and interval")
  ),
   question("True or False: you can add two vectors of different lengths together",
    answer("True",correct = TRUE),
    answer("False")
  ),
  question("In the question above, what is the R 'behavior' that will result from this?",
    answer("Concatinating"),
    answer("Recycling",correct = TRUE),
    answer("An operation"),
    answer("A function")
  ),
  question("If you want to create a vector from 1 to five million in intervals of 1000, what would be the most efficient way to do so in R?",
    answer("c() function"),
    answer(": operator"),
    answer("seq() function",correct = TRUE),
    answer("the function() function")
  
  )
  
)

```

## Dataframes Practice Problems

Q1: Create a dataframe using the `data.frame()` function that is composed of a column/variable called 'value' that is a numeric vector of length 3 ranging from 3 to 27 in intervals of 9 and a column/variable called 'value_2' that is a character vector of length 3 composed of a,b, and c. Name this dataframe df_1.

```{r problem_dataframe, exercise=tutorialize}


df_1 <- data.frame(
  
  value = seq(from = 3,to = 27,by = 9),
  value_2 = c("a","b","c")
  
)



```

Q2: Use the `head()` function to look at the first row of the df_1 dataframe

```{r problem_dataframe_2, exercise=tutorialize,exercise.setup=problem_dataframe}

head(df_1,1)
  
```


Q3: Use the data() function to load in the iris dataframe that comes with R.
```{r problem_dataframe_3, exercise=tutorialize}
data(iris)
```

Q4: Use the head() function to preview the first three rows of the iris dataframe.
```{r problem_dataframe_4, exercise=tutorialize}
head(iris,3)
```

Q5: Use the plot() function and the `~` operator to create an x-y plot of Sepal.Length as a function of Sepal.Width.
```{r problem_dataframe_5, exercise=tutorialize}
plot(Sepal.Length ~ Sepal.Width,data = iris)
```

Q6: Use the plot() function and the `$` operator to create an x-y plot of Sepal.Length as a function of Sepal.Width.
```{r problem_dataframe_6, exercise=tutorialize}
plot(iris$Sepal.Width,iris$Sepal.Length)
```


```{r quiz_dataframes}
quiz(
  question(
    "Which of the characteristics of tidy data is not an atomic vector?",
    answer("a variable/column"),
    answer("a row/observation",correct = TRUE),
    answer("a cell/value")),
  question(
    "What is the R function used to create a data frame from existing vectors?",
    answer("create.df()"),
    answer("df()"),
    answer("data.frame()",correct = TRUE),
    answer("str()")),
  question(
    "What is the R function used to create a frequency distribution of values in a vector, variable, or column?",
    answer("hist()",correct = TRUE),
    answer("boxplot()"),
    answer("plot()"),
    answer("barplot()"))
  
)
```

## Extraction Practice Problems

Q1: Create a numeric vector called new_vec that is a sequence from 500 to 1000 that increases in intervals of 10.
```{r problem_extraction, exercise=tutorialize}
new_vec <- seq(from = 500,to = 1000, by = 10)
```

Q2: Use the `[]` operator to create a another new vector called new_vec_2 that extracts every other value from new_vec starting at 1. Hint: use the `length()` function by writing ?length in the console.
```{r problem_extraction_2, exercise=tutorialize,exercise.setup=problem_extraction}
new_vec_2 <- new_vec[seq(from = 1, to = length(new_vec),by = 2)]
```

Q3: Use the data() function to load in the iris dataframe.
```{r problem_extraction_3, exercise=tutorialize}
data(iris)
```

Q4: Extract the third column and fourth row from the iris dataframe.
```{r problem_extraction_4, exercise=tutorialize}
iris[4,3]
```

Q5: Use the `[]` operator to produce the same output that head(iris,1) would produce.
```{r problem_extraction_5, exercise=tutorialize}
iris[1,]
```

Q6: Use the `$` operator to extract the Petal.Length column.
```{r problem_extraction_6, exercise=tutorialize}
iris$Petal.Width
```
Q7: Use the `mean()` function and the `$` operator to calculate the average Sepal.Length.
```{r problem_extraction_7, exercise=tutorialize}
mean(iris$Sepal.Width)
```

Q7.5: Repeat question 7 but 'wrap' or 'nest' the mean function in the round() function to round the output to two decimal places.
```{r problem_extraction_7_5, exercise=tutorialize}
round(mean(iris$Sepal.Width),2)
```

Q8: Add a new column called 'perc_max_length' to the iris dataframe that calculates each Petal.Length as a percentage of the maximum Petal.Length

```{r problem_extraction_8, exercise=tutorialize}

iris$per_max_length <- (iris$Petal.Length/max(iris$Petal.Length))*100

```

Q9: using the `[]` and `$` operators along with a comparison operator to filter to the row that has the per_max_length value of 100.
```{r problem_extraction_9, exercise=tutorialize,exercise.setup=problem_extraction_8}

iris[iris$per_max_length == 100,]

```

Q10: using the `[]` and `$` operators to filter the iris dataframe to data that is only composed of data from the virginica species
```{r problem_extraction_10, exercise=tutorialize,exercise.setup=problem_extraction_8}

iris[iris$Species == 'virginica',]

```

```{r quiz_extraction}
quiz(
  question(
    "How would you extract the 20th value in a vector called 'secret_data'?",
    answer("secret_data(20)"),
    answer("secret_data[20]",correct = TRUE),
    answer("secret_data.20"),
    answer("secret_data{20}")),
  question(
    "How would you extract the first column vector from the iris dataframe?",
    answer("iris[,1]",correct = TRUE),
    answer("iris(,1)"),
    answer("iris[1]"),
    answer("iris[c(1,)]")),
  quiz(
  question(
    "What is the function of the `$` operator?",
    answer("Add columns",correct = TRUE),
    answer("Extract rows"),
    answer("Extract columns",correct = TRUE),
    answer("Extract from  avector"))
  
),
quiz(
  question(
    "True or False: the `[]` operator only works on vectors",
    answer("True"),
    answer("False",correct = TRUE))
  
),
quiz(
  question(
    "True or False: the `$` operator only works on rows dataframes",
    answer("True"),
    answer("False",correct = TRUE))
  
)
)
```

## Custom Functions Practice Problems

Q1: Create a function called my_range that calculates the differences between the maximum and minimum values in a vector. Hint: use the `min()` and `max()` functions to do so.

```{r problem_function, exercise=tutorialize}

my_range <- function(x){
  
  output <- max(x) - min(x)
  
  return(output)
  
}

```

Q2: Use the my_range() function you calculated to calculate the range of the Petal.Length variable in the iris dataframe.

```{r problem_function_2, exercise=tutorialize,exercise.setup = problem_function}

my_range(iris$Petal.Length)

```
