---
title: "Kashmir"
output: learnr::tutorial
runtime: shiny_prerendered
description: "Using tidy functions across groups in data"
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tutorial.helpers)
library(lterdatasampler)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
```

```{r, include=FALSE}
#Author: Andrew J. Felton
#Date: 11/3/2024
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

```{r copy-code-chunk, child = system.file("child_documents/copy_button.Rmd", package = "tutorial.helpers")}
```

## Summarizing data

Another important data wrangling function in the tidyverse is `summarise()`. Like other tidy functions, the function does as the name says; it summarizes your data! Specifically, it will produce different summary statistics of a variable(s) in your dataframe, with the output being a "summarized" dataframe. A key contrast of `summarise()` with `mutate()` is that `summarise()` collapses the entire dataframe to a one-row dataframe, whereas `mutate()` adds a column and preserves the number of rows in the original dataframe. Indeed, if you use `?` to look at the `summarise()` function's documentation (in your R console), it clearly states that it summarizes each "group" down to one row. We will get to the importance of groups in a moment.

`summarise()` takes on a very similar form as `mutate()`, where we have the dataframe name as the first argument and a name-value pair combination as the second argument. In other words, *to the left of `=` we have the name of the new summarized variable, and to the right of `=` we have the calculation being done*. For example, if you wanted to produce a new dataframe that just contains the average Sepal.Width in the iris dataframe, it would look as such:
```{r demo_1, exercise = TRUE}

#calculate the average Sepal.Width across the entire dataset
dplyr::summarise(iris, average_Sepal.Width = mean(Sepal.Width))

```

The first argument is the dataframe and the second argument is the name-value pair designating the new name of the summarized variable (to the left of `=`) and the summary calculation (to the right of `=`), which will generally entail nesting a summary statistic function, such as `mean()`, within the `summarise()` function. The resulting output is a much smaller dataframe containing the summarized variable of interest.

...And with piping it would look like this:
```{r demo_2, exercise = TRUE}

#calculate the average Sepal.Width across the entire dataset
iris %>% dplyr::summarise(average_Sepal.Width = mean(Sepal.Width))

```

### Q1: Use the `summarise()` function *without* `%>%` to calculate the *minimum* Petal.Width in the iris dataframe, naming the new summarized variable "min_Petal.Width".

```{r problem_1, exercise = TRUE}


```

```{r problem_1-hint}

Use the min() function!

```

```{r problem_1-solution}

iris %>% dplyr::summarise(min_Petal.Width = min(Petal.Width))

```

```{r problem_1-check}

grade_code()

```

### Q2: Use the `summarise()` function *with* `%>%` to calculate the minimum Petal.Width in the iris dataframe, again naming the summarized variable "min_Petal.Width".

```{r problem_2, exercise = TRUE}


```

```{r problem_2-hint}

Where does %>% automatically place the dataframe?

```

```{r problem_2-solution}

iris %>% dplyr::summarise(min_Petal.Width = min(Petal.Width))

```

```{r problem_2-check}

grade_code()

```

You can calculate multiple summary statistics within the `summarise()` function, in the form of adding another name-value pair argument. For example, if you wanted to also calculate the *maximum* Petal.Width along with calculating the minimum, It would look like this:

```{r demo_3, exercise = TRUE}

dplyr::summarise(iris,min_Petal.Width = min(Petal.Width),
                 max_Petal.Width = max(Petal.Width))

```

The above code produce a summarized dataframe with two columns, *yet still one row*. The `summarise()` function is calculating these both summary statistics across the *entire* column, treating the entire dataframe as one "group" to summarize across.

**pro tip**: The `summarise()` function will also work with custom functions. Say you wanted to calculate the range of Petal.Width in addition to its minimum and maximum. Remember how to create a custom function? Let's create one called `range()` that calculates (and *returns*) the range of a variable (or, vector) by subtracting the maximum value from the minimum value:

```{r demo_4, exercise = TRUE}

range <- function(x){
  
  return(max(x) - min(x))
  
}

```

### Q3: Now, use the custom `range()` function to add a third summary statistic in addition to the minimum and maximum Petal.Width that calculates the range of Petal.Width and call this new summarized variable 'range_Petal.Width'. No need for piping here.

```{r problem_3, exercise = TRUE,exercise.setup = "demo_4"}



```

```{r problem_3-hint}

range() gets used just like max() and min() within the summarise() function

```

```{r problem_3-solution}

dplyr::summarise(iris,min_Petal.Width = min(Petal.Width),
                 max_Petal.Width = max(Petal.Width),
                 range_Petal.Width = range(Petal.Width))

```

```{r problem_3-check}

grade_code()

```

The limitation with using `summarise()` on its own is that it will always summarize your dataframe down to one row because it is treating the entire dataframe as one "group". To take your analyses further you need a way to designate distinct groups in your dataframe to summarize across, because often it of more interest to make comparisons among distinct groups (e.g., different species or watersheds or ecosystems) when analyzing environmental data. 

**pro tip**: *`summarize()` will also work (the American way!). I have chosen to be European for this demo and, for the purposes of grading, will suggest you use the European `summarise()`, at least for now.

## Grouping Data

In the prior tutorials, you have been using tidy functions to perform operations across the *entire* dataset. Yet it is often the case that you will want to make comparisons across distinct groups in your dataset. In the penguins dataset, there are distinct species and islands you may wish to make comparisons among. While in the bison dataset there is animal sex, and for in iris dataset there are different species. Species, islands, and animal sex are all examples of variables that you can group your data by. **Grouping data refers to the process of organizing a dataset into subsets based on one or more variables (or columns)**. This allows for easy computation of summary statistics (e.g., the like mean) within each group.

In the tidyverse, there is an important function that helps you designate which variable(s) you want to turn into distinct groups to apply your functions across. This function is called `group_by()` and, like the other tidy functions you have seen, it does as the name it applies; it *groups* your data *by* the unique combinations of values in one or more variables you pass into the function. In the case of grouping by more than one variable, the resulting groups would be all the *unique* combinations of value within the variables passed into the function.

Like most tidy functions, the first argument in `group_by()` is a dataframe, while the second (and so on, designated by `...` in the documentation) argument is the variable(s) to group your data by. You can use many different variables to group the data; what the `group_by()` function will do is group the dataframe (or tibble) according to all the *unique combinations* of the values in the variables specified. 

If you turn back to the iris dataframe, you can use the `distinct()` function to show the different species in the dataset and specifically within the Species column:

```{r demo_5, exercise = TRUE}

iris %>% dplyr::distinct(Species)

```

Okay, great. Now, let's say you want to calculate the average Petal.Length in the iris dataframe. As a first step, it would look something like this:

```{r demo_6, exercise = TRUE}

iris %>% dplyr::summarise(mean_Petal.Length = mean(Petal.Length))

```

Because the iris dataframe is "ungrouped" (the whole dataset is a single group), it summarizes the entire Petal.Length column. However, it might be of interest to calculate this summary statistic for each Species. You can use `group_by()` to group your data across the distinct values within the Species column. It would look like this:

```{r demo_7, exercise = TRUE}

iris %>% 
  dplyr::group_by(Species) %>% #creates a grouped dataframe to pass into summarise()
  dplyr::summarise(mean_Petal.Length = mean(Petal.Length))

```

Because you first grouped the iris dataframe across the distinct Species values, the following `summarise()` function performs the calculation across each distinct group. Underneath the hood, R is "splitting" the dataframe into subsets with each unique group, "applying" the `summarise()` function to each subsetted group, and then combining them back together to one dataframe as the end product.

### Q4: Repeat your workflow in Q3, but this time first group the iris dataframe by Species. Use  `%>%` to do all this within a pipeline.

```{r problem_4, exercise = TRUE}



```

```{r problem_4-hint}

You need to group the data before you summarize it!

```

```{r problem_4-solution}

iris %>% 
  group_by(Species) %>%
  dplyr::summarise(min_Petal.Width = min(Petal.Width),
                 max_Petal.Width = max(Petal.Width),
                 range_Petal.Width = range(Petal.Width))

```

```{r problem_4-check}

grade_code()

```

## Problem Set

For this problem set, you will work with the knz_bison dataset again.

### Q5: First, load the lterdatasampler package. This R package has already been installed for you.
```{r problem_5, exercise=TRUE}


```

```{r problem_5-hint}

Use the library() function!

```

```{r problem_5-solution}

library(lterdatasampler)

```

```{r problem_5-check}

grade_code()

```

### Q6: Create a pipeline that first groups the data by the animal_sex variable *and then* calculates the mean animal_weight, naming the new summarized variable "mean_weight". 

**pro tip**: you will have to use the na.rm=T argument again to tell R to ignore rows with missing values (such is the nature of real data...)!
```{r problem_6, exercise=TRUE}


```

```{r problem_6-hint}

The na.rm=T argument goes inside the mean() function!

```

```{r problem_6-solution}

knz_bison %>% 
  dplyr::group_by(animal_sex) %>%
  dplyr::summarise(mean_weight = mean(animal_weight,na.rm=T))

```

```{r problem_6-check}

grade_code()

```

### Q7: Calculate the average bison age according to sex. Create a pipeline that first calculates a new variable called "age" (as you did in the lab) *and then* groups the data by animal_sex *and then* calculates the average bison age.

```{r problem_7, exercise=TRUE}


```

```{r problem_7-hint}

You will need to use the mutate() function to calculate the age variable

```

```{r problem_7-solution}

knz_bison %>% 
  dplyr::mutate(age = rec_year - animal_yob) %>%
  dplyr::group_by(animal_sex) %>%
  dplyr::summarise(mean_age = mean(age,na.rm=T))

```

```{r problem_7-check}

grade_code()

```

So, female Bison are, on average, much older (and slightly heavier) than male Bison. Interesting! *But do you think this age pattern is consistent for every year?*. To answer this question, you need to group by every unique animal_sex-year combination.

### Q8: Use the `distinct()` and `length()` function to find out how many unique animal_sex-rec_year combinations there are in the knz_bison dataframe. Build a pipeline that first identifies the distinct combinations *and then* summarizes how many there are (by summarizing rec_year). Name the new summarized variable "unique_combos".

```{r problem_8, exercise=TRUE}


```

```{r problem_8-hint}

each variable is an argument in the distinct() function

```

```{r problem_8-solution}

knz_bison %>% 
  dplyr::distinct(rec_year,animal_sex) %>%
  dplyr::summarise(unique_combos = length(rec_year))

```

```{r problem_8-check}

grade_code()

```

*You should have calculated 52 unique combinations.*

### Q9: Now, build a pipeline that first calculates bison age *and then* groups the data by rec_year *and* animal_sex *and then* calculates the average bison age and names this new summarized variable "average_age" *and then* creates an x-y plot with age as the dependent variable (Y axis) and year as the independent variable (X axis).

```{r problem_9, exercise=TRUE}



```

```{r problem_9-hint}

Remember the dot operator

```

```{r problem_9-solution}

knz_bison %>% 
  dplyr::mutate(age = rec_year - animal_yob) %>%
  dplyr::group_by(rec_year,animal_sex) %>%
  dplyr::summarise(average_age = mean(age)) %>%
  plot(average_age ~ rec_year,data = .)

```

```{r problem_9-check}

grade_code()

```

One problem is that the base R `plot()` function does not distinguish between males and females, which is what we are interested in (visually) analyzing. You must do that manually by creating a color palette. Let's use the same approach as in the previous lab, using the `case_when()` function nested in the `mutate()` function to assign males the color "blue" and females the color "red". You got this!

### Q10: Build a pipeline that first calculates bison age *and then* groups the data by rec_year *and* animal_sex *and then* calculates the average bison age and names this new summarized variable 'average_age' *and then* creates a new column called bison_color that designates the color "blue" for males and the color "red" for females *and then* creates an x-y plot with age as the dependent variable (Y axis) and year as the independent variable (X axis).

```{r problem_10, exercise=TRUE}


```

```{r problem_10-hint}

You will need to use the "col" arugment within the plot() function

```

```{r problem_10-solution}

knz_bison %>% 
  dplyr::mutate(age = rec_year - animal_yob) %>%
  dplyr::group_by(rec_year,animal_sex) %>%
  dplyr::summarise(average_age = mean(age)) %>%
  dplyr::mutate(bison_color = case_when(
    animal_sex == "M" ~ "blue",
    animal_sex == 'F' ~ "red")) %>%
  plot(average_age ~ rec_year,data = .,col=bison_color)

```

```{r problem_10-check}

grade_code()

```

### Q11: It looks as though females are consistently older than males through time. Instead of plotting every year, you can compare their ages using a boxplot. Boxplots are useful because they are a way to compare how the distribution of data compares among distinct groups, like male versus female bison. To this end, replace the last x-y plot with a boxplot using the `boxplot()` function, which creates a box-and-whisker plot. Use the `?` operator in the console to find out how to use it!

```{r problem_11, exercise=TRUE}


```

```{r problem_11-hint}

You do not need to color-code for this figure!

```

```{r problem_11-solution}

knz_bison %>% 
  dplyr::mutate(age = rec_year - animal_yob) %>%
  dplyr::group_by(rec_year,animal_sex) %>%
  dplyr::summarise(average_age = mean(age)) %>%
  boxplot(average_age ~ animal_sex,data = .)

```

```{r problem_11-check}

grade_code()

```

Case settled? Do female bison live longer than males?

## Quiz

```{r quiz_1}

quiz(
  question(
    "In R, which function is commonly used to group data by one or more variables?",
    answer("filter()"),
    answer("distinct()"),
    answer("group()"),
    answer("group_by()",correct = TRUE),
    answer("arrange()"))
)
  
```

```{r quiz_2}

quiz(
  question(
    "Based on your analysis in Q's 10-11, what is the most accurate statement?",
    answer("Male bison consistently live longer than females"),
    answer("Female bison consistently live longer than males",correct = TRUE),
    answer("Age difference between male and female bison differ through time"),
    answer("Male and female bison tend to have have near-equal ages"))
  )
  
```

```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```

All done! Please upload your html file in the correct assignment folder.
