---
title: "Vectors and Functions"
output: learnr::tutorial
runtime: shiny_prerendered
description: "Introduction to the use of vectors and functions in R"

---

```{r setup, include=FALSE}
# library(remotes)
# remotes::install_github("rstudio/gradethis")
library(learnr)
library(gradethis)
knitr::opts_chunk$set(echo = FALSE)

# tutorial_event_recorder <- function(tutorial_id, tutorial_version,
#                                     user_id, event, data) {
# cat(tutorial_id, " (", tutorial_version, "): ", user_id , "\n", sep = "")
# cat("event: ", event, "\n", sep = "")
# }
# options(tutorial.event_recorder = tutorial_event_recorder)
# if(interactive()){
#       write_answers(paste0(getwd(), "/", "outfile.pdf"), sess,is_test = F)
# }

```

## Vectors

A `Vector` is a sequences of values, composed of the same `data type`, of a given `length`.

When we say `length`, we mean the number of values in the sequence. The vector 1,2,3 has length = 3 because it has three values in the sequence, and the vector 1,2,3,4 has length = 4 because it has 4 values in the sequence.

When we say `data type`, we mean the type of information a variable is stored as. There are multiple data types that a vector can be stored as. Importantly, a vector is  composed of the *same* data type. We call this an ‘atomic vector’.

*For now, the key data types you will need to know include:*

* numeric: numbers that are either real (double) or integers (integer). These do not have quotations. 

* character: text that come in quotations. Character vectors are composed of strings, which are used to represent text as opposed to numbers. By putting a letter, word, symbol, or number in quotations, you are converting it to a character, or string.

* logical: one of two values: T or F to represent true or false

In R, we can use the `:` operator to rapidly create a numeric sequence. Run the code below to see the output.

```{r sequence_demo, exercise=TRUE}

#create a sequence of 1 to 10 that increases by values of 1
1:10

```

Of course, we can assign this sequence a name, and then it will get stored as that name for later use. Run the code below to see the output.
```{r sequence_demo_2, exercise=TRUE}

#create a sequence of 1 to 10 that increases by values of 1
sequence <- 1:10
sequence 

```

## Functions

Almost everything you do in R will require, to some extent, the use of functions. Indeed, we can only get so far with creating and making use of vectors in the absence of using functions.

A function has three basic characteristics:

* A function has a name that is typically descriptive in some form of what it does.
* Bounded parentheses to the right of the name that contain everything that will be 'passed' into the function. This is in contrast to operators, where whatever is being operated on can be on either side of the symbol - case in point the `:` operator to create a sequence. 
* Arguments that define the parameters that will be fed as inputs to the function, which are needed for the function to return an output. All arguments are contained in the bounded parentheses.

Before going any further run this simple piece of code using the `?` operator.

```{r combine_demo, exercise=TRUE}

#use the help tool to explore the c() function in R
?c

```

The output should be a somewhat weird version of the help tab, but it contains all the essential information. The `c()` function combines (or concatenates) values into a vector. This allows us to create all sorts of vectors. Run the code below to create an (atomic) character vector:

```{r combine_demo_2, exercise=TRUE}

c("R","is","the","best")

```

Notice how each value is separated by  a comma. Commas separate each argument in a function, and here each value is treated as an argument. 


We can also add vectors together if we so wish. Adding two vectors together highlights an important behavior of data in R called `recycling`. If we added two vectors together that had different lengths, what do you think might happen? First let's add two vectors of the same length together. Run the code below:
```{r recyling_demo, exercise=TRUE}

vector_1 <- c(1,1,1,1)
vector_2 <- c(2,2,2,2)

vector_1 + vector_2

```

The above line of code contains a ton of concepts of R already: operators, functions, and vectors. What R does is add the first value of the first vector to the first value of the second vector, and so on. This works well when we have vectors of equal lengths, but what happens when we have vectors of different lengths? Run the code below.

```{r recyling_demo_2, exercise=TRUE}

vector_3 <- c(1,2,3,4)
vector_4 <- c(0,1)

vector_3 + vector_3

```

What happened is called `recycling` and is an important behavior of functions in R, in that the values in vector_4 got "recycled". When it came to find a third value to add to vector_3, vector_4 just went on and repeated itself starting from the beginning back to the first value of the sequence, 0. This is also what is happaneing if we were to simply just add 1 to vector_3; 1 would be added to every value of vector_3 because we are constantly recycling the value of 1.

*Let's look at some other functions we can use the vectors:*

* `seq()`: create a sequence of values in a specific interval
* `mean()`: calculate the mean of a vector
* `plot()`: plot 
* `hist()`: create a histogram or a frequency distribution of values in a vector


```{r functions_demo, exercise=TRUE}

#create a specific sequence of values
seq(from = 1, to = 20, by = 4)

```

Notice the use of the `=` operator. This is how we pass values into functions for a given argument. Notice also how I have specified the names of the arguments. For now this is good practice, but eventually you might realize that R can know which argument you mean by the location of value you are passing into the function.

*it is also good practice to use the `<-` operator specifically for assigning objects names, and the `=` operator for specifying what values will be passed into a function, for a given argument.

Now let's try the `mean()` function to calculate the average of all values in the vector. Run the code below:
```{r functions_demo_2, exercise=TRUE}

#create a specific sequence of values and name it
toy_sequence <- seq(from = 1, to = 20, by = 4)

#calculate the mean of the sequence
mean(toy_sequence)

```

What we did here was pass the *name* of the vector we created (toy_sequence) into the `mean()` function. We can do this because we have stored this vector in our Global Environment, but can reference it as shorthand using the name we have assigned it.

R is also excellent for data visualization, and we can make all sorts of plots using...you guessed it, functions!

Two types of plots we will focus on here:

* Relationship between two variables (an "x-y" plot), using two vectors
* Distribution or frequency of values in one vector

Before proceeding, use the `?` operator to learn more about the `plot()` function.

If we want to see at how two variables correlate with each other, we can use the `plot()` function, which comes standard in the 'base' version of R. Run the code chunk below: 

```{r functions_demo_3, exercise=TRUE}

#create a specific sequence of values and name it
toy_sequence <- seq(from = 1, to = 20, by = 4)

#multiply every value in toy_sequence by 0.5 (this is a type of recycling!)
toy_sequence_halved <- toy_sequence*0.5

#plot the relationship between the two:
plot(x = toy_sequence, y = toy_sequence_halved)

```

The `plot()` function is good for visualizing how two variables correlate with each other: *"how does variable Y change with variable X?*" 

Modify the above code by changing how many values are in the first vector (by varying the interval length in the 'by' argument) and see how things change.

Last but not least, we may also want to visualize the distribution of values in a vector; how frequent is each value? To highlight this, we can generate a toy `normal distribution`, which is the shape of a bell and thus called a bell curve. This is a common distribution the emerges when you have lots of data. For example, if you were to take the height out of every person in the world, a bell-shaped distributed would emerge. We can generate a toy normal distribution using the `rnorm()` function. Use the `?` operator learn more about this function.

```{r functions_demo_4, exercise=TRUE}

#create a vector that is normally distributed, with a length of 20, a mean of 10, and a standard deviation of 2
normal_1 <- rnorm(n = 20,mean = 10, sd = 2)

#plot the distribution of values:
hist(normal_1)

```

That doesn't quite like a nice 'bell' that was promised. What if we increase the length of the vector and thus the 'sample size' to 200 values?

```{r functions_demo_5, exercise=TRUE}

#create a vector that is normally distributed, with a length of 200, a mean of 10, and a standard deviation of 2
normal_2 <- rnorm(n = 200,mean = 10, sd = 2)

#plot the distribution of values:
hist(normal_2)

```

Ah, that is much better. Now let's do some practice questions.

## Problem Set

Use the `:` operator to create a vector that ranges from 10 to 20 in intervals of 1.

```{r vector_practice, exercise=TRUE}

```

```{r vector_practice-hint}

You can do this in one line of code without any functions

```

```{r vector_practice-solution}

10:20

```

```{r vector_practice-check}

grade_this_code()

```

Write the R code to create a vector that ranges from 0 to 100 in intervals of 25.

```{r vector_practice_2, exercise=TRUE}

```

```{r vector_practice_2-hint}

Remember the function needed to create a sequence?

```

```{r vector_practice-2-solution}

seq(from = 0,to = 100,by = 25)

```

```{r vector_practice_2-check}

grade_this_code()

```

Add the R code to the code below that will add alternating values of 0 and 1 to the vector named toy_vector. Use the `?` operator to see what the `rep()` function does. 

```{r vector_practice_3, exercise=TRUE}

zeros <- rep(0,20)

```

```{r vector_practice_3-hint}

Remember how recycling works?

```

```{r vector_practice_3-solution}

zeros <- rep(0,20)
zeros + 0:1

```

```{r vector_practice_3-check}

grade_this_code()

```

Write the R code needed to create a vector named 'class_name' that spells out ENSC311 when you print its name, with each individual letter being a value in the vector. Print the name to see the output.

```{r vector_practice_4,exercise=TRUE}



```

```{r vector_practice_4-hint}

Remember to use the c() function and to use character vectors

```

```{r vector_practice_4-solution}

class_name <- c("E","N","S","C","3","1","1")
class_name

```

```{r vector_practice_4-check}

grade_this_code()

```

Write the R code the needed to create a vector that is normally distributed containing 500 values, has a mean of 5000, and has a standard deviation of 50. Assign this vector the name 'normal_vector' and then create a historgram this vector's distribution of values.


```{r vector_practice_5,exercise=TRUE}



```

```{r vector_practice_5-hint-1}

remember the function you use to create a toy normal distribution of values?

```

```{r vector_practice_5-hint-2}

Remember that you can insert the name of a vector in the hist() function

```

```{r vector_practice_5-solution}

normal_vector <- rnorm(n = 500,mean = 5000,sd = 50)
hist(normal_vector)

```

```{r vector_practice_5-check}

grade_this_code()

```

## Quiz

```{r quiz}

learnr:: quiz(
  question("True or False, all values in an atomic vector must be composed of the same data type",
    answer("True",correct = TRUE),
    answer("False")
  ),
  question("What does the c() function do?",
    answer("prevents a line of code from running"),
    answer("adds two vectors together"),
    answer("combines values in a sequence to create a vector",correct = TRUE),
    answer("creates a sequence of values according to a range and interval")
  ),
   question("True or False, you can add two vectors of different lengths together",
    answer("True",correct = TRUE),
    answer("False")
  ),
  question("In the question above, what concept is this an example of?",
    answer("Concatinating"),
    answer("Recycling",correct = TRUE),
    answer("An operation"),
    answer("A function")
  )
  
  )

```

## Submit

### *Important*: to submit your answers, please generate your hash code, copy it into the decoder, and then decode it into a table. Please copy and save the table in a word document and submit to the D2L assignment for credit. We are actively working to making this last step easier, so thanks for bearing with us!

*Ignore the directions below for submitting to the google link (it doesn't go anywhere)*

```{r context="server"}

learnrhash::encoder_logic()
learnrhash::decoder_logic()
```

```{r encode, echo=FALSE}
learnrhash::encoder_ui()
learnrhash::decoder_ui()
```

All done!
