---
title: "Tidy Functions"
output: learnr::tutorial
runtime: shiny_prerendered
description: "Introducing key tidyverse-based data wrangling functions"
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tutorial.helpers)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
```

```{r, include=FALSE}
#Author: Andrew J. Felton
#Date: 10/18/2024
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

```{r copy-code-chunk, child = system.file("child_documents/copy_button.Rmd", package = "tutorial.helpers")}
```

## The Tidyverse

The `tidyverse` is often contrasted with the `base` R functions and operators we have been working with thus far. "base" refers to the core set of software (e.g., functions) that comes when you initially download R. We can extend the power of R through downloading different *R packages*. The tidyverse is an `R package` which contains a variety of functions for working with data and especially for `data wrangling` and `data visualization`.

-   *Package*: an R package is a collection of functions, datasets, and other code that is bundled together for a specific purpose and thus follows a specific theme. It extends the functionality of the base R software that comes with the initial download of R. One example is the "terra" package, which contains functions to help you work with gridded spatial data.

-   *Data wrangling*: Data wrangling is the process of transforming data from its raw form into another form with the specific intent to make it more valuable and usable for downstream purposes including visualization and statistical analysis. The tidyverse is an R package that has the purpose to extend R's functionality for data wrangling and visualization.

The tidyverse also presents an opportunity to introduce the concept of installing and loading packages in R. Everything you have done in R prior to this can be done with the base R software that comes with the initial download of R. But if you want to extend what you can do with R, you will need to install new packages that contain new functionality. All you need is the name of the package you want to install, and then you can pass that name as a *character* into the `install.packages()` function. If curious, use the `?` operator to learn more about it (it is recommended to do this in the R console). Installing the tidyverse package would look like this: install.packages("tidyverse"). No need to do this here, it has already been done for you; we will revisit this in the near future.

### Important

-   *You only need to install an R package once*.
-   *But, with each new R session, you must load the package*. This is achieved using the `library()` function, where you just pass in the name of the installed package into the function. Since the tidyverse package has already been installed for you, you can load the tidyverse package with the code below:

```{r demo_load, exercise = TRUE}

#load the tidyverse package
library(tidyverse)

```

### Tidy Basics

The tidyverse is built around the idea of organizing data according to "tidy" data principles:

-   Every variable is a column
-   Every observation is a row
-   Every value is a cell

Together, these three principles create a data table (also called "tabular" data). In the tidyverse, dataframes can take the form of `tibbles`, which are just special types of dataframes. The tidy data paradigm is the paradigm within which the tidyverse functions were built. This tutorial is focused on key tidyverse functions that originate from a package called "dplyr" (a riff on "data pliers"). Note that to the left of these functions there will typically be "dplyr::". The `::` operator is used to access specific functions, data, or other objects from a specific package to avoid any conflicts or confusion. This is telling R to take the function being used from that specific package, here the dplyr package, which is actually one of many packages within the broader tidyverse package. Thus, the tidyverse packages in a conglomerate of multiple R packages.

**pro tip:** *The first argument in all of these tidyverse functions is the name of the dataframe!*.

## A note on positional arguments

Up to this tutorial, you have been very explicit in defining the name of the arguments passed into the functions you have been using; these are called `named arguments`. For example:

```{r demo_argument_1, exercise = TRUE}

#named arguments example with iris and head() function
head(x = iris,n = 2)

```

The above is an example of using named arguments by clearly linking the "x" argument (an object) to the iris dataframe and the "n" argument (an integer vector) to the number of rows to preview in the iris dataframe. 

R also can also do `positional arguments` to circumvent naming every argument. Rather, R can infer what the argument is based on the *position* of the value being passed into the function. This is common practice in frequently used functions and especially in tidyverse functions. For example:

```{r demo_argument_2, exercise = TRUE}

#positional arguments example with iris and head() function
head(iris,2)

```

The above is an example of using an example of positional arguments. R understands what each argument is based off its position because in the `head()` function the name of the object always comes *first* while the number of rows to preview always comes *second*.

Tidyverse functions take advantage of R's ability to use positional arguments.

## The `mutate()` function

The `mutate()` function calculates or adds a new variable to a dataframe, **resulting in a new column**. The first (positional) argument passed into the function is the *name of the dataframe*, and the second argument specifies the *new variable*, including the name *and* the value. For example, if you wanted to create and add a new variable to the iris dataframe called "ratio" that is calculated as Sepal.Width divided by Sepal.Length, it would look as such:

```{r demo_mutate, exercise = TRUE}

#create a new variable named ratio using mutate:
iris <- dplyr::mutate(iris,ratio = Sepal.Width/Sepal.Length)
head(iris,1)

```

The second argument in the `mutate()` function is an example of a "name-value" pair; to the left of `=` is the name of the *new variable* while to the right of `=` is the value and/or calculation associated with it. Notice how the calculation above used existing variables.

The analog using an extraction operator would be:

```{r demo_mutate_2, exercise = TRUE}

#create a new variable named ratio using an extraction operator
iris$ratio <- iris$Sepal.Width/iris$Sepal.Length
head(iris,1)
```

The two code chunks above do the same thing and work in the same way. *Remember that variables (columns) in a dataframe are just atomic vectors*. Both approaches above are just dividing two vectors of the same length by each other to generate a new atomic vector which, in this case, produces a new variable/column in the iris dataframe called "ratio".

## The `filter()` function

The `filter()` function excludes (subsets) data according to a set criteria that is specified using a `comparison operator`. The first argument is the name of the dataframe, the second argument is the condition. As one example, you may only be interested in analyzing one species in a dataframe that contains data for multiple species. If you wanted to filter the iris dataframe to include only rows with the setosa species, it would look as such:

```{r demo_filter_1, exercise = TRUE}

#filter iris dataframe to rows with data only for setosa using filter function
setosa_tidy <- dplyr::filter(iris,Species == "setosa")
head(setosa_tidy,5)
```

The analog using extraction operators would be:

```{r demo_filter_2, exercise = TRUE}

#filter iris dataframe to rows with data only for setosa using extraction operator
setosa_extraction <- iris[iris$Species == "setosa",]
head(setosa_extraction,5)

```

## The `select()` function

The `select()` function selects a specific column from the dataframe. The first argument of the function is the name of the dataframe (or tibble), while the second argument is the name (or a vector of names) of the column. There is no need for placing column names in quotations here or even using `c()` to define a sequence of multiple columns. If you wanted to select the Petal.Width column from the iris dataframe, it would look as such:

```{r demo_select_1, exercise = TRUE}

#select Petal.Width column using the select function
petal_width_tidy <- dplyr::select(iris,Petal.Width)
head(petal_width_tidy,5)

```

The analog using an extraction operator would be:

```{r demo_select_2, exercise = TRUE}

#select Petal.Width using an extraction operator
petal_width_extraction <- iris$Petal.Width
head(petal_width_extraction,5)

#could also do this:
petal_width_extraction_2 <- iris[,"Petal.Width"]
head(petal_width_extraction_2,5)

```

Note how the tidy version automatically returns a dataframe, while the extraction operator returns an atomic vector. There are workarounds using operators, but the implicit expectation from tidy is that you'd like to continue working in a dataframe, which is usually the case.

## The `arrange()` function

Finally, you might want to order rows in a dataframe according to ascending or descending values in a variable. The `arrange()` function allows you to do this; the first argument is the name of the dataframe while the second argument is the variable to arrange *all* rows by. The default is ascending but you can specify if you want the order to be descending. For example, if you wanted to order rows in the iris dataframe according to *ascending* values of Petal.Length, it would look as such:

```{r demo_arrange_1, exercise = TRUE}

#arrange rows in the iris dataframe by low to high values of Petal.Length
petal_length_low_high <- dplyr::arrange(iris,Petal.Length)
head(petal_length_low_high,5)

```

If you wanted to order rows according to *descending* values (high to low) of Petal.Length, you need to specify this further using a nested `desc()` function, which you can nest within the `arrange()` function:

```{r demo_arrange_2, exercise = TRUE}

#arrange rows in the iris dataframe by high to low values of Petal.Length
petal_length_high_low <- dplyr::arrange(iris,desc(Petal.Length))
head(petal_length_high_low,5)

```

## Problem Set

### Q1: Mutate it

First, use the `mutate()` function to calculate and add a new variable (column) to the iris dataframe called Petal.Area that is calculated by multiplying Petal.Width by Petal.Length. Make that this new variable gets stored in the iris dataframe.

```{r problem_1, exercise=TRUE}


```

```{r problem_1-hint}

Use the assignment operator to ensure this new variable gets stored in iris

```

```{r problem_1-solution}

iris <- dplyr::mutate(iris,Petal.Area = Petal.Length*Petal.Width)

```

```{r problem_1-check}
grade_code()
```

### Q2: Filter it

Second, use the `filter()` function to filter to rows that are above the *average* value of Petal.Area and assign this filtered dataframe the name "iris_filtered".

```{r problem_2,exercise = TRUE, exercise.setup = "problem_1-solution"}


```

```{r problem_2-hint}

What function calculates an average?

```

```{r problem_2-solution}

iris_filtered <- dplyr::filter(iris,Petal.Area > mean(Petal.Area))

```

```{r problem_2-check}

grade_code()

```

### Q3: Select it

```{r problem_3_setup,include=FALSE}

iris <- dplyr::mutate(iris,Petal.Area = Petal.Length*Petal.Width)
iris_filtered <- dplyr::filter(iris,Petal.Area > mean(Petal.Area))

```

Third, use the `select()` function to return a dataframe with only the Species and Petal.Area columns, using the "iris_filtered" dataframe you created.

```{r problem_3,exercise = TRUE,exercise.setup = "problem_3_setup"}



```

```{r problem_3-hint}

No need to use a nested `c()` function here!

```

```{r problem_3-solution}

dplyr::select(iris_filtered,Species,Petal.Area)

```

```{r problem_3-check}

grade_code()

```

### Q4: Arrange it

```{r problem_4_setup,include=FALSE}

iris <- dplyr::mutate(iris,Petal.Area = Petal.Length*Petal.Width)
iris_filtered <- dplyr::filter(iris,Petal.Area > mean(Petal.Area))

```

Finally, use the `arrange()` function to order the rows of the "iris_filtered" dataframe by descending value of Petal.Area.

```{r problem_4,exercise = TRUE,exercise.setup = "problem_4_setup"}


```

```{r problem_4-hint}

Use a nested `desc()` function!

```

```{r problem_4-solution}

dplyr::arrange(iris_filtered,desc(Petal.Area))

```

```{r problem_4-check}
grade_code()
```

## Quiz

```{r quiz_1}

quiz(
  question(
    "What is the largest value of Petal.Area in iris_filtered?",
    answer("17.97"),
    answer("3.24"),
    answer("15.87",correct = TRUE),
    answer("7.89"))
  
)
  
```

```{r quiz_2}

quiz(
  question(
    "What is always the first argument in dplyr tidy functions",
    answer("The formula"),
    answer("The dataframe name",correct = TRUE),
    answer("The variable name"),
    answer("The comparison"))
  
  )

```

```{r quiz_3}
  
quiz(
  question(
    "What tidy function corresponds to this code: data[data&variable > 2,]?",
    answer("filter()",correct = TRUE),
    answer("arrange()"),
    answer("mutate()"),
    answer("select()"))
)
    
```

```{r quiz_4}

quiz(
    question(
    "What tidy function corresponds to this code: data$variable_3 <- data$variable_1 + data$variable_2?",
    answer("filter()"),
    answer("arrange()"),
    answer("mutate()",correct = TRUE),
    answer("select()"))
)

```

```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```

All done! Please upload your html file in the correct assignment folder.
