---
title: "Operators"
output: learnr::tutorial
runtime: shiny_prerendered
description: "Introduction to using operaters in R"
---

```{r setup, include=FALSE}
# library(remotes)
# remotes::install_github("rstudio/gradethis")
library(learnr)
library(gradethis)
library(learnrhash)
knitr::opts_chunk$set(echo = FALSE)
gradethis::gradethis_setup()
options(tutorial.event_recorder = tutorialize::recorder)
options(tutorial.storage = learnr::filesystem_storage(file.path(path.expand("~"), "_learnr_storage_")))
```

## Operators and names 

`Operators` are symbols that perform logical or mathematical manipulations (or operations) on values or variables in data. Operators that perform calculations, such as addition, are `arithmetic operators'. The symbols for these are intuitive:

*Arithmetic operators:*

* `+` does addition
* `-` does subtraction
* `*` does multiplication
* `/` does division
* `^` Does 'to the power of` (e.g., to square or cube a value)

In R, we often will want to `assign` a value (and eventually a dataset!) a `name`. This is useful because once we do this, we can store the value in the `Global Environment`, allowing us to reference this value for later use. We can use an operator called the `assignment operator` to do this.

*Assignment operator:*

* `<-` assigns a value a name

You can think of this as an arrow, pointing to what name a value corresponds to. Pay attention to the sequence of this code:

*If we run this code, R will return the value 75. Go ahead and run the code!*
```{r demo, exercise=TRUE}

75

```

*We can `assign` this value a name. Typically we want this name to be descriptive of the value or have some 'syntactic sugar'. Run the code and see what (if any) the output is:*
```{r demo_assign, exercise=TRUE}

seventy_five <- 75

```

*We can also can add more syntactic sugar by `commenting`our data using `#`. Anything following this symbol will not run as code, but will be visible, and is typically used to describe what you are doing.*
```{r demo_comment, exercise=TRUE}

#I am creating a variable called seventy_five
seventy_five <- 75

```

*Notice that R returns nothing when we run the above code. What is happening here is that R has stored the value 75 with the name seventy_five. Now we can reference this name to return the value 75. Naming objects is thus a way we can store them. Run the code and see what happens:*
```{r demo_comment_return, exercise=TRUE}

#I am creating a variable called seventy_five
seventy_five <- 75

#return the value
seventy_five

```

Modify the above blocks of code with different names and values to see for yourself.

## Problem Set

*Let's start out by using arithmetic operators to do some calculations*

Write the R code required to add two plus two.

```{r problem_two-plus-two, exercise=TRUE}

```

```{r problem_two-plus-two-hint-1}

This is just like using a calculator!

```

```{r problem_two-plus-two-solution}

2 + 2

```

```{r problem_two-plus-two-check}

grade_this_code()

```

Write the R code required to calculate 10 divided by 5.

```{r problem_divide, exercise=TRUE}



```

```{r problem_divide-hint}

Use the slash symbol

```

```{r problem_divide-solution}

10/5

```

```{r problem_divide-check}

grade_this_code()

```

*Now we want to be able to assign a name to the outputs of our calculations so we can store them*

Assign the value of your previous calculation to the name `two`, then print the output.

```{r problem_name-divide, exercise=TRUE}



```

```{r problem_name-divide-hint-1}

 Can you remember the special operator we use to assign names to objects?

```

```{r problem_name-divide-solution}
two <- 10/5
two
```

```{r problem_name-divide-check}

grade_this_code()

```

Create another named objected called twenty_five that is 5 squared.

```{r problem_name-square, exercise=TRUE}



```

```{r problem_name-square-hint}

Use the `^` operator 

```

```{r problem_name-square-solution}

twenty_five <- 5^2

```

```{r problem_name-square-check}

grade_this_code()

```

*Named objects can interact with each other. Let's calculate a third object.*

Finally, multiply `two` by `twenty_five` and name this calculation as a third value called `fifty`. Print it so that the new object returns the value of the calculation. For now, you will have to 're-assign' each value its name, as tutorials do not carry over this type of information across different code chunks.

```{r problem_name-multiply, exercise=TRUE}



```

```{r problem_name-multiply-hint-1}

Because the named objects are just numbers, they will behave as such

```

```{r problem_name-multiply-solution}

two <- 10/5
twenty_five <- 5^2
fifty <- two*twenty_five
fifty

```

```{r problem_name-multiply-check}

grade_this_code()

```

## Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
learnr::quiz(
  question("When you assign a name to an object, where does that object get stored?",
    answer("In a Script"),
    answer("In the Console"),
    answer("In the Global Environment", correct = TRUE),
    answer("In the Help/Plots")
  ),
  question("What is the purpose of the `<-` operator?",
    answer("to prevent a line of code from running"),
    answer("add two values"),
    answer("compare two values"),
    answer("assign a name to a value", correct = TRUE)
  )
)
```

## Submit

### *Important*: to submit your answers, please press `generate` to create you unique code, copy the code into the box under 'hash to decode', and then press `decode` to turn your code it into a table summarizing your tutorial answers. Please copy the table in a word document, save it, and submit it to the D2L Operators assignment for credit. We are actively working on making this last submission step better and easier, so thanks for bearing with us!

*Ignore the directions below for submitting to the google link (it doesn't go anywhere)!*

```{r context="server"}

learnrhash::encoder_logic()
learnrhash::decoder_logic()
```

```{r encode, echo=FALSE}
learnrhash::encoder_ui()
learnrhash::decoder_ui()
```

All done! Please submit your table in a word document to the D2L assignment.
