---
title: "Smooth Operator"
output: learnr::tutorial
runtime: shiny_prerendered
description: "Exploring the concept and use of operators in R"
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tutorial.helpers)
knitr::opts_chunk$set(echo = FALSE)
```

```{r, include=FALSE}
#Author: Andrew J. Felton
#Date: 9/18/2024
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

```{r copy-code-chunk, child = system.file("child_documents/copy_button.Rmd", package = "tutorial.helpers")}
```

## Operators 

`Operators` are symbols that perform logical (e.g., true or false) or mathematical (e.g., subtraction or addition) manipulations (also called "operations") on values or variables in data. Operators that perform calculations are called `arithmetic operators`. The symbols for these are (mostly) intuitive:

*Arithmetic operators:*

* `+` does addition
* `-` does subtraction
* `*` does multiplication
* `/` does division
* `^` Does "to the power of" (e.g., to square or cube a value)

In R, you can also `assign` objects (whether a number, vector, or dataset) a `name`. This is useful because once we do this, we can store the `objects` in the `Global Environment`. This ability to store objects in R is important because it will allow you to reference them for later use in your analysis. You can use an operator called the `assignment operator` to give objects names and store them in the global environment for later use, referencing the name assigned to it. The utility of storing named objects in the global environment will become more evident when you starting using R scripts.

*Assignment operator:*

* `<-` assigns an object a name

You can think of this as an arrow, pointing to what the name of an object is. Pay attention to the sequence of this code:

*If you run this code, R will return the value 75. Go ahead and run the code!*
```{r demo, exercise=TRUE}

75

```

*You can `assign` this number, 75, a name. Typically, you will want this name to be descriptive or informative and thus have some `syntactic sugar`; coding features that make the code easier or more readable (e.g., descriptive and concise object names) but don't necessarily increase code functionality or performance. Run the code and see what the output is (or is not):*
```{r demo_assign, exercise=TRUE}

seventy_five <- 75

```

* `Pro tip:` When naming objects, do not use spaces (R will not like this). Instead, use underscores. For example, instead of naming an object "seventy five", you should name it "seventy_five".

*You can also can add more syntactic sugar by `commenting` your code using the `#` operator. Anything following # will not run as code, but will be visible to the reader of your code. This is useful for two reasons: 1) to describe and annotate what your code is doing or 2) to `comment out` a line or block of code so that is doesn't run.*
```{r demo_comment, exercise=TRUE}

#I am creating a variable called seventy_five
seventy_five <- 75

```

*Notice that R returns nothing when we run the above code. What is happening here is that R has stored the value 75 with the name seventy_five. Now we can reference this name to return the value 75. Naming objects is thus a way you can store them in R's memory for later use. Run the code and see what happens:*
```{r demo_comment_return, exercise=TRUE}

#I am creating a variable called seventy_five
seventy_five <- 75

#print the value of the object names seventy_five
seventy_five

```

Modify the above blocks of code with different names and values to see for yourself.

## Problem Set 

*Let's start by using arithmetic operators to do some calculations*

Write the R code required to add two plus two.

```{r problem_1, exercise=TRUE}


```

```{r problem_1-hint-1}

This is just like using a calculator!

```

```{r problem_1-solution}

2 + 2

```

```{r problem_1-check}

grade_this_code()

```

Write the R code required to calculate 10 divided by 5.

```{r problem_2, exercise=TRUE}



```

```{r problem_2-hint}

Use the slash symbol

```

```{r problem_2-solution}

10/5

```

```{r problem_2-check}

grade_this_code()

```

Assign the value of your previous calculation the name `two`, then print the output.

```{r problem_3, exercise=TRUE}



```

```{r problem_3-hint-1}

 Can you remember the special operator we use to assign names to objects?

```

```{r problem_3-solution}
two <- 10/5
two
```

```{r problem_3-check}

grade_this_code()

```

Create another named object called twenty_five that is 5 squared.

```{r problem_4, exercise=TRUE}



```

```{r problem_4-hint}

Use the `^` operator 

```

```{r problem_4-solution}

twenty_five <- 5^2

```

```{r problem_4-check}

grade_this_code()

```

*Named objects can interact with each other. Calculate a third object.*

Multiply `two` by `twenty_five` and name the result of calculation a third object called `fifty`. Print it so that the new object returns the value of the calculation. 

```{r problem_5-setup, include=FALSE}

twenty_five <- 5^2
two <- 10/5

```


```{r problem_5, exercise=TRUE,exercise.setup = "problem_5-setup"}



```

```{r problem_5-hint-1}

Because the named objects are just numbers, they will behave as such

```

```{r problem_5-solution}

fifty <- two*twenty_five
fifty

```

```{r problem_5-check}

grade_this_code()

```

## Quiz

```{r quiz}
learnr::quiz(
  question("When you assign a name to an object, where does that object get stored?",
    answer("In a Script"),
    answer("In the Console"),
    answer("In the Global Environment", correct = TRUE),
    answer("In the Help/Plots")
  ),
  question("What is the purpose of the `<-` operator?",
    answer("to prevent a line of code from running"),
    answer("add two values"),
    answer("compare two values"),
    answer("assign a name to an object", correct = TRUE)
  ),
  question("What is an example of an object in R?",
    answer("a dataset"),
    answer("a vector"),
    answer("an equation"),
    answer("all the above", correct = TRUE)
    )
)
```


```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```

All done! Please upload your html file in the correct assignment folder.

