---
title: "Operators"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
# library(remotes)
# remotes::install_github("rstudio/gradethis")
library(learnr)
library(gradethis)
knitr::opts_chunk$set(echo = FALSE)
```

## Operators and names 

`Operators` are symbols that perform logical or mathematical manipulations (or operations' on values or variables in data. Operators that perform calculations, such as addition, are `arithmetic operators'. The symbols for these are intuitive:

*Arithmetic operators:*

* `+` does addition
* `-` does subtraction
* `*` does multiplication
* `/` does division
* `^` Does 'to the power of` (e.g., to square or cube a value)

In R, we often will want to `assign` a value (and eventually a dataset!) a `name`. This is useful because once we do this, we can store the value in the `Global Environment`, allowing us to reference this value for later use. We can use an operator called the `assignment operator` to do this.

*Assignment operator:*

* `<-` assigns a value a name

You can think of this as an arrow, pointing to what name a value corresponds to. Pay attention to the sequence of this code:

*If we run this code, R will return the value 75:*
```{r demo, exercise=TRUE}

75

```

*We can `assign` this value a name. Typically we want this name to be descriptive of the value or have some 'syntactic sugar':*
```{r demo_assign, exercise=TRUE}

seventy_five <- 75

```

*We can also can add more sugar by `commenting`our data using `#`. Anything following this symbol will not run as code, but will be visible, and is typically used to describe what you are doing.*
```{r demo_comment, exercise=TRUE}

#I am creating a variable called seventy_five
seventy_five <- 75

```

*Notice R returns nothing when we run the above code. What is happening here is that R has stored the value 75 with the name seventy_five. Now we can reference this name to return the value 75 (run the code and see what happens). Naming objects is a way we can store them:*
```{r demo_comment_return, exercise=TRUE}

#I am creating a variable called seventy_five
seventy_five <- 75

#return the value
seventy_five

```

Modify the above blocks of code to see for yourself.

## Problem Set

*Let's start out by using arithmetic operators to do some calculations*

Write the R code required to add two plus two

```{r two-plus-two, exercise=TRUE}

```

```{r two-plus-two-hint-1}

This is just like using a calculator!

```

Write the R code required to calculate 10 divided by 5

```{r divide, exercise=TRUE}

```

*Now we want to be able to assign a name to the outputs of our calculations so we can store them*

Assign the value of your previous calculation to the name `two`, then print the output:

```{r name-divide, exercise=TRUE}



```

```{r name-divide-hint-1}

 Can you remember the special operator we use to assign names to objects?

```

```{r name-divide-solution}
two <- 10/5
two
```

```{r name-divide-check}

grade_this_code()

```

Create another named objected called twenty_five that is 5 squared

```{r name-square, exercise=TRUE}



```

```{r name-square-hint}

Use the `^` operator 

```

```{r name-square-solution}

twenty_five <- 5^2

```

```{r name-square-check}

grade_this_code()

```

*Named objects can interact with each other. Calculate a third object*

Finally, multiply `two` by `twenty_five` and name this calculation
`fifty`. Print so that the new object returns the value of the calculation.

```{r name-multiply, exercise=TRUE}



```

```{r name-multiply-hint}

Because the named objects are just numbers, they will behave as such

```

```{r name-multiply-solution}

fifty <- five*twenty_five
fifty

```

```{r name-multiply-check}

grade_this_code()

```

## Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("When you assign a name to an object, where does that object get stored?",
    answer("In a Script"),
    answer("In the Console"),
    answer("In the Global Environment", correct = TRUE),
    answer("In the Help/Plots")
  ),
  question("What is the purpose of the `<-` operator?",
    answer("to prevent a line of code from running"),
    answer("add two values"),
    answer("compare two values"),
    answer("assign a name to a value", correct = TRUE)
  )
)
```
