---
title: "Extraction"
tutorial:
  id: "Week_5_Extraction"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
description: "Basics of extracting and adding data to a dataframe"
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
gradethis::gradethis_setup()
options(tutorial.event_recorder = tutorialize::recorder)
options(tutorial.storage = learnr::filesystem_storage(file.path(path.expand("~"), "_learnr_storage_")))
```

## Extraction Basics

A core part of data exploration and analysis is the ability to subset your data into smaller units. Key to this is the idea of data *extraction*, which focuses on retrieving specific subsets of your data. These subsets may be a single or multiple values that meet a specific criteria of interest. You may additionally want to extract entire columns (variables) from a dataframe. You can further use extraction to add new data to your dataframe.

There are two key operators useful for extracting data from a dataframe:

* The `$` operator allows you to extract or *add* entire columns (variables).
* The `[]` operator allows you to extract specific values based on a reference to a row, column, or both.

## Extracting data

Because dataframes are just special data structures composed of different atomic vectors (columns), the `[]` operator works on vectors. Run the code below to create a toy vector

```{r demo_extraction, exercise = TRUE}

#create and print a vector
toy_vector <- seq(from = 10, to = 100, by = 10)
toy_vector

```

We can use the `[]` operator to *extract* values from vectors. Let's say we want to extract the third value in toy_vector. Run the code below.

```{r demo_extraction_2, exercise = TRUE,exercise.setup='demo_extraction'}

#extract third value from the vector
toy_vector[3]

```

Notice two things. First, the operator comes *after* the object it is operating on. Second, the location of the value we are interested in is specified via a numeric value inside the brackets. This extraction is not restricted to one value; we can actually extract multiple values using a vector. Run the code below

```{r demo_extraction_2_5, exercise = TRUE,exercise.setup='demo_extraction'}

#extract multople values from the vector
toy_vector[c(3,5,7)]

```

Now let's say we are creating a dataframe. When we do this, we are moving from a one-dimensional data structure (vector) to a two-dimensional data structure. In other words, we now have columns *and* rows. The `[]` operator allows us to find specific rows *and* columns in a dataframe. In this way, we can extract from more than just atomic vectors.

First, let's create a toy dataframe. Run the code below.

```{r demo_extraction_3, exercise = TRUE}

#define vectors
mean_C_uptake <-  c(700,400,200)
biome <- c("forest",'grassland','desert')

#create dataframe from vectors
biome_data <- data.frame(mean_C_uptake,biome)

#print dataframe
biome_data

```

Now let's say we want to *extract* the values from the third row in the second column of the biome_data dataframe we have created. We can use this `[]` operator do this. Run the code below.

```{r demo_extraction_4, exercise = TRUE,exercise.setup='demo_extraction_3'}

#extract the value from the third row and second column
biome_data[3,2]

```

The first value inside the brackets always refers to the row number, whereas the second value inside the brackets always refers to the column number. By providing the row and column number, the `[]` can map this to a specific value in the biome_data dataframe.

*In the code chunk below, play around with what happens if you add [2,] or [2] or [,2] to the end of biome_data; what do you notice?*

```{r demo_extraction_5, exercise = TRUE,exercise.setup='demo_extraction_3'}

#play around with what value go in the brackets
biome_data[]

```

The `$` operator is particularly good if we want to extract or reference specific columns (variables). This operator gets 'sandwiched' in between the dataframe name and column name we are referencing. Run the example code below.

```{r demo_extraction_6, exercise = TRUE,exercise.setup='demo_extraction_3'}

#extract the biome column
biome_data$biome

```

Change the value of biome to mean_C_uptake in the code chunk above to see what happens.

Note that the `$` operator is less flexible than the `[]` operator and is specific to using with dataframes. However, the `$` operator can be quite useful if we want to quickly calculate a summary statistic on a particular variable. Consider the `max()` function, which just calculates (or really, finds) the maximum values across a numeric vector. If we wanted to find the maximum carbon uptake in the biome_data dataframe, it would like like this.

```{r demo_extraction_6_5, exercise = TRUE,exercise.setup='demo_extraction_3'}

#find the maximum carbon uptake
max(biome_data$mean_C_uptake)

```

Why does this work? It works because we are simply performing a calculation (using a function) on an atomic vector, albeit one that has been *extracted* from a dataframe. To reaffirm this, we can use the `is.atomic()` function on the extracted to column to assess whether or not it is 'true' that what is being passed into the function is an atomic vector. The function returns a logical TRUE or FALSE. Run the code below.

```{r demo_extraction_6_75, exercise = TRUE,exercise.setup='demo_extraction_3'}

#determine whether the extracted column is an atomic vector
is.atomic(biome_data$mean_C_uptake)

```

Looks like its 'true'!

## Adding new variables to a dataframe

In addition to extracting a column (variable) from a dataframe, we can also use the `$` operator to add new columns to a dataframe.

Let's go back to the biome_data dataframe we created in the previous section. Below is a refresher of what that looks like; run the code below.

```{r demo_extraction_7, exercise = TRUE,exercise.setup='demo_extraction_3'}

 #view the dataframe
 head(biome_data)

```

Now let's say we want to add a column that will provide additional information pertaining to what type of plants are dominant within each biome. We can use the `$` operator to do so. Run the code below.
```{r demo_extraction_8, exercise = TRUE,exercise.setup = 'demo_extraction_3'}

#add a plant variable
biome_data$dominant_plants <- c("tree","grass","shrub")

#print the dataframe
biome_data

```

Often times we will want to calculate a new variable using existing variables. We can do this by using vector math and, specifically, by using recycling. Let's say we want to convert the mean_C_uptake column from grams (the units it is in) to kg. We can do this by 1) following the basic approach of the previous code chunk and 2) simply multiplying the mean_C_uptake column by 0.001 (which, in essence, is a vector of 1~). Run the code below

```{r demo_dataframe_9, exercise = TRUE,exercise.setup = "demo_extraction_8"}

#convert from grams to kg
biome_data$mean_C_uptake_kg <- biome_data$mean_C_uptake*0.001

#print dataframe
biome_data
```

Calculating new variables from existing variable can also include equations that include using multiple variables - for example if we wanted to calculate the difference between or ratio of two numeric variables.

## Filter data by a condition

Finally, we can bring the `[]` and `$` operators together to filter data. Filtering data involves excluding data according to a set criteria. In R, there are `comparison operators` that can help define relationships or filtering criteria and use that to exclude data that doesn't meet that criteria.

* `==`	Denotes complete equality (exact match). For example,	x == y.
* `!=`	Denotes 'is not equal to'. For example,	x != y.
* `>` Denotes	'greater than'. For example,	x > y.
* `<`	Denotes 'less than'. For example,	x < y.
* `>=` Denotes 'Greater than or equal to'. For example,	x >= y.
* `<=` Denotes 'Less than or equal to'. For example,	x <= y.

We can use these comparison operators in addition to our extraction operators to filter data by certain criteria. For example, if we wanted to filter to just the row of data with the forest biome, it would look like this. Run the code below.

```{r demo_extraction_10, exercise=TRUE,exercise.setup='demo_extraction_8'}

#filter dataframe to just the forest group
biome_data[biome_data$biome == 'forest', ]

```

Beginning from the inside out, we first extracted the entire biome column using the `$` operator. We then designated that we are looking for an *exact* match via the `==` operator. All of this is occurring within the brackets of the `[]` operator, which if you remember has both a rows and column designation for extracting from dataframes. Because we have our filtering criteria only for rows, it will return the entire row that meets out criteria, along with all columns (since no criteria was set). Think about why it would be most useful to return all rows (and thus a smaller version of the initial dataframe) instead of just a single column or a single value.

*Note that it is often the case that we will want to create a new, 'filtered' dataframe from another dataframe. To do this, you will just need to assign the filtered dataframe a name using the `<-` operator.*

in the same way, we can filter according to a numeric criteria. 
```{r demo_extraction_11, exercise=TRUE,exercise.setup='demo_extraction_8'}

#filter dataframe to just the maximum 
biome_data[biome_data$mean_C_uptake > 200, ]

```
All rows of data that have mean_C_uptake greater than 200 are returned.

A final operator we can use is the `%in%` operator, which assess whether elements in one vector are present in another, and will return a logical TRUE or FALSE whether or not that is the case. We can, however, use it to quickly filter data according to a set criteria. 
```{r demo_extraction_12, exercise=TRUE,exercise.setup='demo_extraction_8'}

#filter dataframe to rows with grasslands and desert biomes
biome_data[biome_data$biome %in% c('grassland','desert'),]

```

Now on to some practice questions!

## Problem set

### Q1: Use the `[]` operator to extract the third,seventh,10th,13th, 256th,1000th, 3000th, and 3500th values from the atomic vector below. First, run the code chunk below first containing this vector.

code <- seq(from = 300,to = 3000, by = 0.5)

```{r new, exercise = TRUE,exercise = tutorialize}

code <- seq(from = 300,to = 3000, by = 0.5)

```

```{r problem_extraction_2, exercise = tutorialize,exercise.setup=new}
code[c(3,7,10,13,256,1000,3000,3500)]
```

### Q2: Now we are going to work with another built-in R dataset called iris. This dataset contains information about different leaf traits for different species. Use the ? operator to learn more about these data. In the code chunk below, run data(iris) to load in the dataframe.

```{r problem_extraction_3, exercise = tutorialize}
data(iris)
```

Q2a: Use the 'head()` function to look at the first 3 rows of the iris dataframe
```{r problem_extraction_4, exercise = tutorialize,exercise.setup=problem_extraction_3}
head(iris,3)
```

Q2b: Use the the `$` operator and the mean() function to calculate the average Petal.Width in the iris dataframe.

```{r problem_extraction_5, exercise = tutorialize,exercise.setup=problem_extraction_4}
mean(iris$Petal.Width)
```

*Can you figure out how to round the above answer to two decimal points? Hint: run ?round() in the console.*

Q2c: Use the `$` operator to add a new column (variable) into the iris dataframe that is calculated as the ratio of Petal.Length length to Petal.Width. Name this new variable 'ratio' and wrap your calculation in the `round()` function to round it to two decimal points.
```{r problem_extraction_6, exercise = tutorialize,exercise.setup=problem_extraction_4}
iris$ratio <- round(iris$Sepal.Length/iris$Petal.Width,2)
```

Q2d: Calculate the range of the ratio variable, which is the maximum minus the minum value of ratio. Hint, use the `min()` and `max()` functions.

```{r problem_extraction_7, exercise = tutorialize,exercise.setup=problem_extraction_6}
max(iris$ratio) - min(iris$ratio)
```

### Q3 Finally, filter the iris dataframe to rows of data that are above the average of all ratio values.
```{r problem_extraction_8, exercise = tutorialize,exercise.setup=problem_extraction_6}

iris[iris$ratio > mean(iris$ratio),]

```

Bonus: can you use the same basic approach as in Q3 to find which species has the maximum ratio?

```{r problem_extraction_9, exercise = tutorialize,exercise.setup=problem_extraction_6}

iris[iris$ratio == max(iris$ratio),]

```

## Quiz 

```{r quiz_tidy}
quiz(
  question(
    "How would you extract the fifth value in a vector called 'secret_data'?",
    answer("secret_data(5)"),
    answer("secret_data[5]",correct = TRUE),
    answer("secret_data.5"),
    answer("secret_data{5}"))
)
```

```{r quiz_dataframe}
quiz(
  question(
    "How would you extract the third column and fifth row from the iris dataframe?",
    answer("iris[5,3]",correct = TRUE),
    answer("iris(5,3)"),
    answer("iris[3,5]"),
    answer("iris[c(3,5)]"))
  
)
```

```{r quiz_head}
quiz(
  question(
    "What is the function of the `$` operator?",
    answer("Add columns",correct = TRUE),
    answer("Extract rows"),
    answer("Extract columns",correct = TRUE),
    answer("Extract from  avector"))
  
)
```

## Submitting tutorial results

Great job!  You've completed the assignment for Lecture 6!  Now it's time to turn it in...

In the console, enter the following (case sensitive):

```{r, echo=TRUE, eval=FALSE}
ENSC311::submit_ENSC311()
```

Your should see something like this:

```{r}
"Week_5_Extraction_311_<Posit_Cloud_ID>"
```

where the end of the character string is an ID that Posit Cloud assigned you. As you do more tutorials, this list of tutorials you have run will get longer. You must pass the number that corresponds to this assignments. For example, *if* you wanted to turn in the first item on the list, you would enter: ENSC311::submit_ENSC311(1).

A "submit_ENSC311_..." file will appear under the "Files" tab in the lower right pane of RStudio.  Check the box next to the submit file and choose "Export..." from the "More" dropdown on the bar at the top of the tab.  Download the file and sumbit it to D2L.

That's it!  You're all done!

