---
title: "When the Levee Breaks"
tutorial:
  id: "Week_6_Exam_Review"
output: 
  learnr::tutorial:
    progressive: true
runtime: shiny_prerendered
description: "Practice problems for Exam II focused on the basics of R"
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tutorial.helpers)
knitr::opts_chunk$set(echo = FALSE)
```

```{r, include=FALSE}
#Author: Andrew J. Felton
#Date: 12/3/2024
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

```{r copy-code-chunk, child = system.file("child_documents/copy_button.Rmd", package = "tutorial.helpers")}
```

## Key Concepts to Review for Exam 2

-   **Atomic Vectors**: be able to define, create, and use them.
-   **Recycling**: be able to define, identify, and apply this concept to data.
-   **Operators** including `:`, `<-`, `$`, `[]`, and all arithmetic and comparison operators.
-   **Functions** including `c()`,`plot()`,`seq()`,`head()`,`str()`,`mean()`, and others covered through the extraction tutorial assignment. *Be able to write a simple custom function.*
-   **Dataframes** including what they are, the three tidy principles, how to make them, and how to manipulate them.
-   **Extraction**: be able to define, identify, and apply this concept to data using operators.

**Pro tip**: In addition to the practice problems below, the best way to study for the exam is to review each tutorial and lab assigned throughout this section.

## Vectors Practice Problems

### Q1: Use the `:` operator to create a numeric vector from 1 to 10 that increases by values of 1. Assign this vector the name "vector_1"

```{r problem_1, exercise=TRUE}


```

```{r problem_1-hint}

Using a function to do this is unnecessary  

```

```{r problem_1-solution,exercise.reveal_solution = FALSE}

vector_1 <- 1:10

```

```{r problem_1-check}

grade_code()

```

### Q2: Use the `c()` function to create a character vector that spells out BIOME, with each value being its own letter. Assign this vector the name "biome".

```{r problem_2, exercise=TRUE}


```

```{r problem_2-hint}

"Use quotations"

```

```{r problem_2-solution,exercise.reveal_solution = FALSE}

biome <- c("B","I","O","M","E")

```

```{r problem_2-check}

grade_code()

```

### Q3: Use recycling to multiply vector_1 by 0.75.

```{r problem_3, exercise=TRUE,exercise.setup="problem_1-solution"}


```

```{r problem_3-hint}

Remember 0.75 is a vector of length 1

```

```{r problem_3-solution,exercise.reveal_solution = FALSE}

vector_1*0.75

```

```{r problem_3-check}

grade_code()

```

### Q4: Create a new vector using the `seq()` function which ranges from 10 to 80 and increases by values of 10. Assign this new vector the name "vector_2". Then, on the next line use the `plot()` function to plot vector_1 (created in Q1 on the x axis and vector_2 on the y axis.

```{r problem_4, exercise=TRUE,exercise.setup="problem_1-solution"}


```

```{r problem_4-hint}

You can pass each vector into the plot() function as an argument.

```

```{r problem_4-solution,exercise.reveal_solution = FALSE}

vector_2 <- seq(from = 8,to = 80,by = 8)
plot(vector_1,vector_2)

```

```{r problem_4-check}

grade_code()

```

### Multiple Choice Questions

```{r quiz_1}

learnr::quiz( 
  question("What is the purpose of the `<-` operator?",
    answer("to prevent a line of code from running"),
    answer("add two values"),
    answer("compare two values"),
    answer("assign a name to an object", correct = TRUE))
  )

```
  
```{r quiz_2} 

learnr::quiz(
  question("What is the purpose of the `:` operator?",
           answer("indicate complete equality"),
           answer("create a sequence of numbers",correct = TRUE),
           answer("compare two values"),
           answer("extract a value from a sequence"))
  )
```

```{r quiz_3}

learnr::quiz(
  question("True or False: all values in an atomic vector are composed of the same data type",
           answer("True",correct = TRUE),
           answer("False"))
  )

```

```{r quiz_4}

learnr::quiz(
  question("What does the c() function do?",
           answer("prevents a line of code from running"),
    answer("adds two vectors together"),
    answer("combines values in a sequence to create a vector",correct = TRUE),
    answer("creates a sequence of values according to a range and interval"))
)

```

```{r quiz_5}
  
learnr::quiz(
  question("True or False: you can add two vectors of different lengths together",
    answer("True",correct = TRUE),
    answer("False"))
)
  
```

```{r quiz_6}

learnr::quiz(
  question("What is the question above an example of?",
    answer("Concatinating"),
    answer("Recycling",correct = TRUE),
    answer("An operation"),
    answer("A function"))
  )

```
  
```{r quiz_7}

learnr::quiz(
  question("If you want to create a vector from 1 to five million in intervals of 1000, what would be the most efficient way to do so in R?",
    answer("c() function"),
    answer(": operator"),
    answer("seq() function",correct = TRUE),
    answer("the function() function"))
  )

```

## Dataframes Practice Problems

### Q5: Create a dataframe using the `data.frame()` function that is composed of 2 variables: 1) a column/variable called "value" that is a numeric vector of length 3 ranging from 3 to 27 in intervals of 9 and 2) a column/variable called "value_2" that is a character vector of length 3 composed of the value a,b, and c (in that order). Assign this dataframe the name "df_1".

```{r problem_5, exercise=TRUE}

df_1 <- data.frame(
  
  value = seq(from = 3,to = 27,by = 9),
  value_2 = c("a","b","c")
  
)

```

```{r problem_5-hint}

The creation of each column is an argument within the data.frame() function 

```

```{r problem_5-solution,exercise.reveal_solution = FALSE}

df_1 <- data.frame(
  
  value = seq(from = 3,to = 27,by = 9),
  value_2 = c("a","b","c")
  
)

```

```{r problem_5-check}

grade_code()

```

### Q6: Use the `head()` function to look at the first row of the df_1 dataframe

```{r problem_6, exercise=TRUE,exercise.setup="problem_5"}


```

```{r problem_6-hint}

The second argument is the number of rows to reveal
  
```

```{r problem_6-solution,exercise.reveal_solution = FALSE}

head(x = df_1,n = 1)
  
```

```{r problem_6-check}

grade_code()

```

### Q7: Use the `data()` function to load in the iris dataframe that comes with R.

```{r problem_7, exercise=TRUE}


```

```{r problem_7-hint}

There is only ONE argument here!
  
```

```{r problem_7-solution,exercise.reveal_solution = FALSE}

data(iris)
  
```

```{r problem_7-check}

grade_code()

```

### Q8: Use the `head()` function to preview the first three rows of the iris dataframe.

```{r problem_8, exercise=TRUE}


```

```{r problem_8-hint}

The first argument is the name of the data.frame

```

```{r problem_8-solution,exercise.reveal_solution = FALSE}

head(x = iris,n = 3)

```

```{r problem_8-check}

grade_code()

```

### Q9: Use the `plot()` function with the `~` operator to create an x-y plot of Sepal.Length as *a response to* Sepal.Width.

```{r problem_9, exercise=TRUE}


```

```{r problem_9-hint}

Remember that the "response" variable goes on the Y axis of an X-Y plot

```

```{r problem_9-solution,exercise.reveal_solution = FALSE}

plot(Sepal.Length ~ Sepal.Width,data = iris)

```

```{r problem_9-check}

grade_code()

```

### Q10: Use the `plot()` function with the `$` operator to create an x-y plot of Sepal.Length as *a response to* Sepal.Width.

```{r problem_10, exercise=TRUE}


```

```{r problem_10-hint}

With `$`, each variable is a seperate argument int he `plot()` function

```

```{r problem_10-solution,exercise.reveal_solution = FALSE}

plot(iris$Sepal.Width,iris$Sepal.Length)

```

```{r problem_10-check}

grade_code()

```

### Multiple Choice Questions

```{r quiz_8}

learnr::quiz(
  question(
    "Which of the characteristics of tidy data is not an atomic vector?",
    answer("a variable/column"),
    answer("a row/observation",correct = TRUE),
    answer("a cell/value"))
  )

```

```{r quiz_9}

learnr::quiz(
  question(
    "What is the R function used to create a data frame from existing vectors?",
    answer("create.df()"),
    answer("df()"),
    answer("data.frame()",correct = TRUE),
    answer("str()"))
  )

```

```{quiz_10}

learnr::quiz(
  question(
    "What is the R function used to create a frequency distribution of values in a vector, variable, or column?",
    answer("hist()",correct = TRUE),
    answer("boxplot()"),
    answer("plot()"),
    answer("barplot()"))
)


```

## Extraction Practice Problems

### Q11: Create a numeric vector that is a sequence from 500 to 1000 that increases in intervals of 10. Assign this vector the name "new_vec".

```{r problem_11, exercise=TRUE}


```

```{r problem_11-hint}

Which function helps you create a sequence with specific intervals?

```

```{r problem_11-solution,exercise.reveal_solution = FALSE}

new_vec <- seq(from = 500,to = 1000, by = 10)

```

```{r problem_11-check}

grade_code()

```

### Q12: Use the `[]` operator to create a numeric vector that is the result of extracting every other value from "new_vec", starting at its first value. Assign this new vector the name "new_vec_2".

**Pro tip**: Remember to use the help operator `?` in the R console to learn more about how a specific function works (e.g., ?head).

```{r problem_12, exercise=TRUE,exercise.setup="problem_11-solution"}


```

```{r problem_12-hint}

Use the `length()` function 

```

```{r problem_12-solution,exercise.reveal_solution = FALSE}

new_vec_2 <- new_vec[seq(from = 1, to = length(new_vec),by = 2)] 

```

```{r problem_12-check}

grade_code()

```

### Q13: Use the `data()` function to load in the iris dataframe.

```{r problem_13, exercise=TRUE}


```

```{r problem_13-hint}

You already did this earlier!
  
```

```{r problem_13-solution,exercise.reveal_solution = FALSE}

data(iris)
  
```

```{r problem_13-check}

grade_code()

```

### Q14: Extract the third column and fourth row from the iris dataframe.

```{r problem_14, exercise=TRUE}


```

```{r problem_14-hint}

Row-Column

```

```{r problem_14-solution,exercise.reveal_solution = FALSE}

iris[4,3]

```

```{r problem_14-check}

grade_code()

```

### Q15: Use the `[]` operator to produce the same output that head(iris,1) would produce.

```{r problem_15, exercise=TRUE}


```

```{r problem_15-hint}

You may not need to worry about specifying columns here

```

```{r problem_15-solution,exercise.reveal_solution = FALSE}

iris[1,]

```

```{r problem_15-check}

grade_code()

```

### Q16: Use the `$` operator to extract the Petal.Length column.

```{r problem_16, exercise=TRUE}


```

```{r problem_16-hint}

Instead of a number you use the name of the column/variable with `$`.

```

```{r problem_16-solution,exercise.reveal_solution = FALSE}

iris$Petal.Length

```

```{r problem_16-check}

grade_code()

```

### Q17: Use the `mean()` function and the `$` operator to calculate the average Sepal.Length.

```{r problem_17, exercise=TRUE}


```

```{r problem_17-hint}

The answer to question 17 is now an argument in the `mean()` function

```

```{r problem_17-solution,exercise.reveal_solution = FALSE}

mean(iris$Sepal.Length)

```

```{r problem_17-check}

grade_code()

```

### Q18: Repeat question 17 but "wrap" or "nest" the `mean()` function as an argument *inside* the `round()` function so to round the output to two decimal places.

```{r problem_18, exercise=TRUE}


```

```{r problem_18-hint}

The `round()` functions hould have two arguments in it

```

```{r problem_18-solution,exercise.reveal_solution = FALSE}

round(mean(iris$Sepal.Length),2)

```

```{r problem_18-check}

grade_code()

```

### Q19: Use `$` to add a new column named "perc_max_length" to the iris dataframe that calculates each Petal.Length as a percentage of the maximum Petal.Length.

```{r problem_19,exercise=TRUE}


```

```{r problem_19-hint}

This will involve using a mix of operators and the `max()` function

```

```{r problem_19-solution,exercise.reveal_solution = FALSE}

iris$perc_max_length <- (iris$Petal.Length/max(iris$Petal.Length))*100

```

```{r problem_19-check}

grade_code()

```

### Q20: Using the `[]` and `$` operators along with a comparison operator to filter to the row(s) that has a "perc_max_length value" of 100.

```{r problem_20, exercise=TRUE,exercise.setup = "problem_19"}


```

```{r problem_20-hint}

Remember you are just filtering by row

```

```{r problem_20-solution,exercise.reveal_solution = FALSE}

iris[iris$perc_max_length == 100,]

```

```{r problem_20-check}

grade_code()

```

### Q21: Using the `[]` and `$` operators to filter the iris dataframe to data that is only composed of data from the virginica species

```{r problem_21, exercise=TRUE,exercise.setup="problem_19-solution"}


```

```{r problem_21-hint}

Remember you are just filtering by rows with this species

```

```{r problem_21-soluton,exercise.reveal_solution = FALSE}

iris[iris$Species == 'virginica',]

```

```{r problem_21-check}

grade_code()

```

### Multiple Choice Questions

```{r quiz_11}

learnr::quiz(
  question(
    "How would you extract the 20th value in a vector named secret_data?",
    answer("secret_data(20)"),
    answer("secret_data[20]",correct = TRUE),
    answer("secret_data.20"),
    answer("secret_data{20}"))
)
  
```

```{r quiz_12}

learnr::quiz(
  question(
    "How would you extract the first column from the iris dataframe?",
    answer("iris[,1]",correct = TRUE),
    answer("iris(,1)"),
    answer("iris[1]"),
    answer("iris[c(1,)]"))
  )

```

```{r quiz_13}

learnr::quiz(
  question(
    "What is the function of the `$` operator?",
    answer("Add columns",correct = TRUE),
    answer("Extract rows"),
    answer("Extract columns",correct = TRUE),
    answer("Extract from  avector"))
)

```
  
```{r quiz_14}
 
learnr::quiz( 
  question(
    "True or False: the `[]` operator only works on vectors",
    answer("True"),
    answer("False",correct = TRUE))
)

```
  
```{r quiz_15}

learnr::quiz(
  question(
    "True or False: the `$` operator only works on rows dataframes",
    answer("True"),
    answer("False",correct = TRUE))
)

```

## Custom Functions Practice Problems

### Q22: Create a function named my_range that calculates the differences between the maximum and minimum values in a vector.

```{r problem_22, exercise=TRUE}


```

```{r problem_22-hint}

Use the `min()` and `max()` functions in your custom function.

```

```{r problem_22-solution,exercise.reveal_solution = FALSE}

my_range <- function(x){
  
 max(x) - min(x)
  
}

```

```{r problem_22-check}

grade_code()

```

### Q23: Use the `my_range()` function you created to calculate the range of the Petal.Length variable within the iris dataframe.

```{r problem_23, exercise=TRUE,exercise.setup = "problem_22-solution"}

my_range(iris$Petal.Length)

```

```{r problem_23-hint}

The column (or really, vector) should be the sole argument in the function

```

```{r problem_23-solution,exercise.reveal_solution = FALSE}

my_range(iris$Petal.Length)

```

```{r problem_23-check}

grade_code()

```

```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```

All done! Please upload your html file in the correct assignment folder.
