---
title: "One more time"
output: 
  learnr::tutorial:
    progressive: true
tutorial:
  id: "Week_5_Dataframes"
runtime: shiny_prerendered
description: "Introducing the basics of dataframes in R"
---
```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tutorial.helpers)
knitr::opts_chunk$set(echo = FALSE)
```

```{r, include=FALSE}
#Author: Andrew J. Felton
#Date: 9/24/2024
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

```{r copy-code-chunk, child = system.file("child_documents/copy_button.Rmd", package = "tutorial.helpers")}
```

## Data frame basics

The data table is the primary data structure you will use to store and analyze data. In base R, this is realized through the data frame. Data frames are a core data structure in R, providing a *tabular* (or, *table-based*) structure for organizing, storing, and analyzing your data; as such, they are R's original version of a data table. This tutorial will guide you through the fundamentals of data frames in R.

Often, the best way to structure data frames is by following the `tidy data` approach. Do you remember the three "rules" of tidy data?

* Every *column* is a *variable*.
* Every *row* is an *observation*.
* Every *cell* is a *value*.

Think for a second about what a column in a data frame really is. If we were to look at one variable (or column) in a tidy data frame, what we are really doing is looking at an atomic vector, or, a sequence of values composed of the same data type. Think about the tabular data you have seen so far; is a column ever composed of different data types?

How about a row in a data frame? Is this an atomic vector?

```{r quiz_1}
quiz(
  question(
    "Is a row in a data frame an atomic vector?",
    answer("Yes"),
    answer("No",correct = TRUE)
))
```

## Building a data frame

In reference to the quiz question, a row (observation) is not *always* an atomic vector, because the sequence of values across a row can be composed of different data types (think of categories versus numbers). Depending on the dataset, a row might turn out to be an atomic vector, but for the purposes of this class **a column is always an atomic vector, but a row is not**. *Ultimately, what this means is that tidy-based data frames are just combinations of different atomic vectors defined by their columns*. By definition, this means that a data frame is a type of `recursive vector`, a `list`. But in effect, a data frame stands on its own as a data structure and is arguably the primary data structure you will use.

Let's belabor the point even more. You can create a data frame from scratch in R using a (you guessed it) *function* called `data.frame()`. The fundamental argument in the `data.frame()` function is one or more vectors, which define the variables and thus columns of the data frame. Run the code below.

```{r demo_dataframe, exercise = TRUE}

#create a data frame using two vectors as arguments
toy_data <- data.frame(
  
  var_1 =  1:10,
  var_2 = seq(from = 10, to = 100, by = 10)
  
)

toy_data

``` 

Notice that in the above data frame that the code returns, it provides the *type* of data each variable is being stored as below the column (variable) name: "int" means an *integer* or whole number (no decimals), while "dbl" means *double* or a real number (decimals). We can create the same data frame in a different way by first creating and assigning the vectors names outside of the function and then just passing their names into the `data.frame()` function as arguments. Run the code below.

```{r demo_dataframe_2, exercise = TRUE}

#create a data frame using two vectors already assigned names
var_1 <-  1:10
var_2 <- seq(from = 10, to = 100, by = 10)

toy_data <- data.frame(var_1,var_2)

toy_data

```

Both variables in the above data frame are numbers. We can add character vectors into this to create a categorical variable in a data frame. Run the code below to create a hypothetical data frame containing data on carbon uptake and biome type.

```{r demo_dataframe_3, exercise = TRUE}

#create a data frame with both numeric and character vectors
mean_C_uptake <-  c(700,400,200)
biome <- c("forest",'grassland','desert')

biome_data <- data.frame(mean_C_uptake,biome)

biome_data

```

Notice how the data type for the biome variable is "chr", which means *character*, and refers to data stored as text (strings).

## Plotting and viewing data frames

When you create or import data frames you will typically want to preview them. A key function that allows us to do so is the `head()` function, which previews the "head" or "top" of the data frame by looking at the first X number of rows (while retaining the column header/variable names).

Let's say you have created a data frame. Run the code below.
```{r demo_dataframe_4, exercise = TRUE}

#create a data frame using two numeric vectors
var_1 <-  1:10
var_2 <- seq(from = 10, to = 100, by = 10)

toy_data <- data.frame(var_1,var_2)

```

You might want to take a quick look at this data frame before working with it. You can use the `head()` function to *preview* the first **n** number of rows. This function accepts a data frame,"x", and a number of rows, "n", as its arguments. Run the code below to see for yourself.

```{r demo_dataframe_5, exercise = TRUE,exercise.setup = "demo_dataframe_4"}

#look at the first two rows of the data frame
head(x = toy_data,n = 2)

```

If you remember, we can plot two vectors together:

```{r demo_dataframe_6, exercise = TRUE}

#create a specific sequence of values and name it
toy_sequence <- seq(from = 1, to = 20, by = 4)

#multiply every value in toy_sequence by 0.5 (this is a type of recycling!)
toy_sequence_halved <- toy_sequence*0.5

#plot the two vectors
plot(x = toy_sequence, y = toy_sequence_halved)

```

We can then combine the two vectors into a data frame, and plot them as well, though in a slightly different way when drawing from a data frame:

```{r demo_dataframe_7, exercise = TRUE,exercise.setup = "demo_dataframe_6"}

#combine the two vectors we previously created into a data frame:
toy_data_2 <- data.frame(vector_1 = toy_sequence,
                         vector_2 = toy_sequence_halved)

#plot vector_1 as a response to vector_2
plot(vector_1 ~ vector_2,data = toy_data_2)
```
Notice how the arguments in the plot function have changed slightly to look more like a formula. Indeed, the `~` operator separates the right and left side of a formula, and is used in many statistical models in R but can also be used with the `plot()` function to create an x-y plot. To the left of `~` is the *response variable*, which will be depicted on the Y axis, while to the right is the *predictor variable*, which will be depicted on the X axis.

## Problem set

### Q1: Create a data frame that has two columns. One column should be called `Letter` that is a character vector spelling out ENSC311, with a letter for each element of the vector. The other column should be called `Number` and should be a numeric vector ranging from 1 to 7, increasing by values of 1, and thus have a length of 7. Assign this data frame the name `ENSC311`.
```{r problem_1, exercise=TRUE}

```

```{r problem_1-hint}

Do not create the atomic vectors outside the data.frame() function, 
just do it all within, treating the vectors (columns) as arguments.

```

```{r problem_1-solution}

ENSC311 <- data.frame(
  
  Letter = c('E','N','S','C','3','1','1'),
  Number = 1:7
  
)

```

```{r problem_1-check}
grade_code()
```

### Q2: Below is code that will load a built-in data frame of the waiting time between eruptions and the duration of the eruption for the Old Faithful geyser in Yellowstone National Park, Wyoming, USA. The dataset is called called 'faithful'. R has this dataset (and many others) stored in its base software, which can be accessed by using the `data()` function and passing the name of the dataset into the function as the argument. Run the code below to load in the dataset.

```{r setup_2, exercise = TRUE}

data(faithful)

```

Q2A: Use the `head()` function to preview the first three rows of the faithful data frame.

```{r problem_2, exercise=TRUE,exercise.setup="setup_2"}

```

```{r problem_2-hint}

Remember this function has two arguments: the name of the data frame and the number of rows.

```

```{r problem_2-solution}

head(x = faithful,n = 3)

```

```{r problem_2-check}
grade_code()
```

Q2B: Use the `plot()` function and your understanding of the variables in the faithful data frame to make an x-y plot that will answer this question: *How does the time between eruptions (the predictor variable) influence the duration of eruptions (the response variable)?*

```{r problem_3, exercise=TRUE}

```

```{r problem_3-hint}

Remember which axis correponds to the response and predictor variable and how the position of each variable relative to `~` influences which axis it goes on?

```

```{r problem_3-solution}

plot(eruptions ~ waiting,data = faithful)

```

```{r problem_3-check}
grade_code()
```

## Quiz 

```{r quiz_2}

#R does not like that there are two answers - need to revise this moving forward
quiz(
  question(
    "Which two characteristics of tidy data are always an atomic vector?",
    answer("a variable and value",correct = TRUE),
    answer("a variable and observation"),
    answer("a value and observation"),
    answer("none of the above"))
)
```

```{r quiz_3}
quiz(
  question(
    "What is the R function used to create a data frame from atomic vectors?",
    answer("create.df()"),
    answer("df()"),
    answer("data.frame()",correct = TRUE),
    answer("str()"))
  
)
```

```{r quiz_4}
quiz(
  question(
    "What code would you write to view the first three rows of a data frame called df?",
    answer("view(n = df,x = 3)"),
    answer("summary(x = df,n = 3)"),
    answer("data.frame(x = df,x = 3)"),
    answer("head(x = df,n = 3)",correct = TRUE))
  
)
```

```{r quiz_5}
quiz(
  question(
    "How does the time between Old Faithful eruptions influence the eruption duration?",
    answer("No clear effect"),
    answer("A longer waiting time leads to a longer eruption",correct = TRUE),
    answer("A longer waiting time leads to a shorter eruption"))
  
)
```

```{r quiz_6}
quiz(
  question(
    "True or false: A data frame is a list",
    answer("True",correct = TRUE),
    answer("False"))
  
)
```

* **pro tip**: `view()`, `summary()`, and `str()` are all real R functions that are also great for understanding a data frame. Here is a short description of each:

* `view()` allows you to see the *entire* data frame
* `summary()` provides a broader overview of the variables and basic *summary statistics* of them in a data frame
* `str()` provides information on the *structure* of an R object, including its dimensions, the type of data structure,variables, and data types.

```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```

All done! Please upload your html file in the correct assignment folder.
