---
title: "One more time"
output: 
  learnr::tutorial:
    progressive: true
tutorial:
  id: "Week_5_Dataframes"
runtime: shiny_prerendered
description: "Introducing the basics of dataframes in R"
---
```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tutorial.helpers)
knitr::opts_chunk$set(echo = FALSE)
```

```{r, include=FALSE}
#Author: Andrew J. Felton
#Date: 9/24/2024
```

```{r info-section, child = system.file("child_documents/info_section.Rmd", package = "tutorial.helpers")}
```

```{r copy-code-chunk, child = system.file("child_documents/copy_button.Rmd", package = "tutorial.helpers")}
```

## Dataframe Basics

Dataframes are a core data structure in R, providing a *tabular* (or, *table-based*) structure for organizing, storing, and analyzing your data; as such, they are also sometimes referred to as "data tables". This tutorial will guide you through the fundamentals of dataframes in R.

Often, the best way to structure dataframes is by following the `tidy data` approach. Do you remember the three "rules" of tidy data?

* Every *column* is a *variable*.
* Every *observation* is a *row*.
* Every *value* is a *cell*.

Think for a second about what a column in a dataframe really is. If we were to look at one variable (or column) in a tidy data frame, what we are really doing is looking at an atomic vector, or, a sequence of values composed of the same data type. Think about the tabular data you have seen so far; is a column ever composed of different data types?

How about a row in a data frame? Is this an atomic vector?

```{r quiz_row}
quiz(
  question(
    "Is a row in a dataframe always an atomic vector?",
    answer("Yes"),
    answer("No",correct = TRUE)
))
```

## Building a dataframe

In reference to the quiz question, a row (observation) is not *always* an atomic vector, because the sequence of values across a row can be composed of different data types (think of categories versus numbers). Depending on the dataset, a row might turn out to be an atomic vector, but for the purposes of this class **a column is always an atomic vector, but a row is not**. *Ultimately, what this means is that tidy-based dataframes are just combinations of atomic vectors defined by their columns*.

Let's belabor the point even more. You can create a dataframe from scratch in R using a (you guessed it) *function* called `data.frame()`. The fundamental argument in the `data.frame()` function is one or more vectors, which define the variables and thus columns of the dataframe. Run the code below.

```{r demo_dataframe, exercise = TRUE}

#create a dataframe using two vectors as arguments
toy_data <- data.frame(
  
  var_1 =  1:10,
  var_2 = seq(from = 10, to = 100, by = 10)
  
)

toy_data

``` 

Notice that in the above dataframe that the code returns, it provides the *type* of data each variable is being stored as below the column (variable) name: "int" means an *integer* or whole number (no decimals), while "dbl" means *double* or a real number (decimals). We can create the same dataframe in a different way by first creating and assigning the vectors names outside of the function and then just passing their names into the `data.frame()` function as arguments. Run the code below.

```{r demo_dataframe_2, exercise = TRUE}

#create a dataframe using two vectors already assigned names
var_1 <-  1:10
var_2 <- seq(from = 10, to = 100, by = 10)

toy_data <- data.frame(var_1,var_2)

toy_data

```

Both variables in the above dataframe are numbers. We can add character vectors into this to create a categorical variable in a dataframe. Run the code below.

```{r demo_dataframe_3, exercise = TRUE}

#create a dataframe with both numeric and character vectors
mean_C_uptake <-  c(700,400,200)
biome <- c("forest",'grassland','desert')

biome_data <- data.frame(mean_C_uptake,biome)

biome_data

```

Notice how the data type for the biome variable is "chr", which means *character*, and refers to data stored as text (strings).

## Plotting and viewing dataframes

When you create or import dataframes you will typically want to preview them first. A key function that allows us to do so is the `head()` function, which allows you to preview the "head" or "top" of the dataframe by looking at the first X number of rows (while retaining the column header/variable names).

Let's say you have created a dataframe. Run the code below.
```{r demo_dataframe_4, exercise = TRUE}

#create a dataframe using two numeric vectors
var_1 <-  1:10
var_2 <- seq(from = 10, to = 100, by = 10)

toy_data <- data.frame(var_1,var_2)

```

You might want to take a quick look at this dataframe before working with it. You can use the `head()` function to *preview* the first **n** number of rows. This function accepts a dataframe,"x", and a number of rows, "n", as its arguments. Run the code below to see for yourself.

```{r demo_dataframe_5, exercise = TRUE,exercise.setup = "demo_dataframe_4"}

#look at the first two rows of the dataframe
head(x = toy_data,n = 2)

```

If you remember, we can plot two vectors together:

```{r demo_dataframe_6, exercise = TRUE}

#create a specific sequence of values and name it
toy_sequence <- seq(from = 1, to = 20, by = 4)

#multiply every value in toy_sequence by 0.5 (this is a type of recycling!)
toy_sequence_halved <- toy_sequence*0.5

#plot the two vectors
plot(x = toy_sequence, y = toy_sequence_halved)

```

We can then combine the two vectors into a dataframe, and plot them as well, though in a slightly different way when drawing from a dataframe:

```{r demo_dataframe_7, exercise = TRUE,exercise.setup = "demo_dataframe_6"}

#combine the two vectors we previously created into a dataframe:
toy_data_2 <- data.frame(vector_1 = toy_sequence,
                         vector_2 = toy_sequence_halved)

#plot vector_1 as a response to vector_2
plot(vector_1 ~ vector_2,data = toy_data_2)
```
Notice how the arguments in the plot function have changed slightly to look more like a formula. Indeed, the `~` operator separates the right and left side of a formula, and is used in many statistical models in R but can also be used with the `plot()` function to create an x-y plot. To the left of `~` is the *response variable*, which will be depicted on the Y axis, while to the right is the *predictor variable*, which will be depicted on the X axis.

## Problem set

### Q1: Create a dataframe that has two columns. One column should be called `Letter` that is a character vector spelling out ENSC311, with a letter for each value. The other column should be called `Number` and should be a numeric vector ranging from 1 to 7, increasing by values of 1, and thus have a length of 7. Assign this dataframe the name `ENSC311`.
```{r problem_1, exercise=TRUE}

```

```{r problem_1-hint}

Remember how to create an atomic vector?

```

```{r problem_1-solution}

ENSC311 <- data.frame(
  
  Letter = c('E','N','S','C','3','1','1'),
  Number = 1:7
  
)

```

```{r problem_1-check}
grade_code()
```

### Q2: Below is code that will load a built-in dataset of the waiting time between eruptions and the duration of the eruption for the Old Faithful geyser in Yellowstone National Park, Wyoming, USA. The dataset is called called 'faithful'. R has this dataset (and many others) stored in its base software, which can be accessed by using the `data()` function and passing the name of the dataset into the function as the argument. Run the code below to load in the dataset.

```{r setup_problem_2, exercise = TRUE}

data(faithful)

```

Q2A: Use the `head()` function to preview the first three rows of the faithful dataframe.

```{r problem_2, exercise=TRUE,exercise.setup="setup_problem_2"}

```

```{r problem_2-hint}

Remember this function has two arguments: the name of the dataframe and the number of rows.

```

```{r problem_2-solution}

head(x = faithful,n = 3)

```

```{r problem_2-check}
grade_code()
```

Q2B: Use the `plot()` function and your understanding of the variables in the faithful dataframe to make an x-y plot that will answer this question: *How does the time between eruptions (the predictor variable) influence the duration of eruptions (the response variable)?*

```{r problem_3, exercise=TRUE}

```

```{r problem_3-hint}

Remember which axis correponds to the response and predictor variable and how the position of each variable relative to `~` influences which axis it goes on?

```

```{r problem_3-solution}

plot(eruptions ~ waiting,data = faithful)

```

```{r problem_3-check}
grade_code()
```

## Quiz 

```{r quiz_1}
quiz(
  question(
    "Which of two characteristics of tidy data are always an atomic vector?",
    answer("a variable/column",correct = TRUE),
    answer("a row/observation"),
    answer("a cell/value",correct = TRUE),
    answer("none"))
)
```

```{r quiz_2}
quiz(
  question(
    "What is the R function used to create a data frame from atomic vectors?",
    answer("create.df()"),
    answer("df()"),
    answer("data.frame()",correct = TRUE),
    answer("str()"))
  
)
```

```{r quiz_3}
quiz(
  question(
    "What code would you write to view the first three rows of a data frame called df?",
    answer("view(n = df,x = 3)"),
    answer("summary(x = df,n = 3)"),
    answer("data.frame(x = df,x = 3)"),
    answer("head(x = df,n = 3)",correct = TRUE))
  
)
```

```{r quiz_4}
quiz(
  question(
    "How does the time between Old Faithful eruptions influence the eruption duration?",
    answer("No clear effect"),
    answer("A longer waiting time leads to a longer eruption",correct = TRUE),
    answer("A longer waiting time leads to a shorter eruption"))
  
)
```

* **pro tip**: `view()`, `summary()`, and `str()` are all real R functions that are also great for understanding a dataframe. Here is a short description of each:

* `view()` allows you to see the *entire* dataframe
* `summary` provides a broader overview of the variables and basic *summary statistics* of them in a dataframe
* `str()` provides information on the *structure* of an R object, including its dimensions, the type of data structure,variables, and data types.

```{r download-answers, child = system.file("child_documents/download_answers.Rmd", package = "tutorial.helpers")}
```

All done! Please upload your html file in the correct assignment folder.
